---
title: "CCRIX_GBRS_Testing"
author: "Dawson Budke"
date: "2025-04-08"
---

## 1. Load in Files

```{r}
# Clear environment to start fresh (if necessary)
rm(list = ls())

#UNC/Jackson labs genotyping microarray consensus file
CC_diplotype_probs <- readRDS("Data/CC_diplotype_probs.rds")

#Anno for above 
anno = read.csv("Data/gm_uwisc_v4.csv")
# check mini experiment below: v4 including markers annotated to bp_grcm39, which was used in my GBRS samples, so its the superior annotation here 

# Grcm39 v105 Ensembl anno 

```

## 2. Filter CC_diplotypes files

```{r}
# Function to filter CC_diplotype_probs to only include specified strains
filter_diplotype_probs <- function(gprob, strain_patterns) {
  # Create a filtered copy of the genoprob object
  filtered_gprob <- gprob
  
  # Get all sample names from the first chromosome (they should be the same across all chromosomes)
  all_samples <- dimnames(gprob[[1]])[[1]]
  
  # Find samples matching the patterns
  matching_indices <- integer(0)
  matching_samples <- character(0)
  
  for (pattern in strain_patterns) {
    matches <- grep(pattern, all_samples, ignore.case = TRUE)
    if (length(matches) > 0) {
      matching_indices <- c(matching_indices, matches)
      matching_samples <- c(matching_samples, all_samples[matches])
    }
  }
  
  # Check if we found any matches
  if (length(matching_indices) == 0) {
    stop("No samples found matching the specified patterns")
  }
  
  # Print the matched samples for verification
  cat("Found", length(matching_samples), "samples matching your criteria:\n")
  cat(paste(matching_samples, collapse = "\n"), "\n")
  
  # Filter each chromosome to keep only the matched samples
  for (chr_name in names(gprob)) {
    filtered_gprob[[chr_name]] <- gprob[[chr_name]][matching_indices, , , drop = FALSE]
  }
  
  # Return the filtered object with all attributes preserved
  attributes(filtered_gprob) <- attributes(gprob)
  return(filtered_gprob)
}

# Use the function with your strain patterns
strain_patterns <- c("CC001/Unc", "CC019/TauUnc", "CC040/TauUnc")
filtered_CC_diplotype_probs <- filter_diplotype_probs(CC_diplotype_probs, strain_patterns)

# Check the dimensions of the first chromosome in the filtered data
dim(filtered_CC_diplotype_probs[[1]])

# You can also verify the sample names in the filtered data
filtered_samples <- dimnames(filtered_CC_diplotype_probs[[1]])[[1]]
print(filtered_samples)

# Save the filtered data if needed

```

## 3. Load in Metadata

```{r}
# CCRIX samples (176) 

# Load in Excel file and read the "pups" sheet
pup <- read.xlsx("Data/Data4R_db_03102025_FINAL.xlsx", sheet = "Pups", na.strings = c(""," ","NA","N/A","na", "n/a"), skipEmptyRows = TRUE)

# Add Cross column 
pup$Cross <- paste0(pup$CC.Mom.Strain, "x", pup$CC.Dad.Strain)

#Edit Low Protein to LP and AIN-93G to AIN (removing space/- since it messes up contrasts function)
pup$Diet <- gsub("Low Protein", "LP", pup$Diet)
pup$Diet <- gsub("AIN-93G", "AIN", pup$Diet)

# Select only the required columns

```

```{r}
# Control Pups (70 samples)

# Load in Excel file and read the "ControlPups" sheet
ControlPup <- read.xlsx("Data/Data4R_db_03102025_FINAL.xlsx", sheet = "ControlPups", na.strings = c(""," ","NA","N/A","na", "n/a"), skipEmptyRows = TRUE)

# Add Cross column 
ControlPup$Cross <- paste0(ControlPup$CC.Mom.Strain, "x", ControlPup$CC.Dad.Strain)

#Edit Low Protein to LP and AIN-93G to AIN (removing space/- since it messes up contrasts function)
ControlPup$Diet <- gsub("Low Protein", "LP", ControlPup$Diet)
ControlPup$Diet <- gsub("AIN-93G", "AIN", ControlPup$Diet)

# Select only the required columns

```

## 4. Interpolate markers to gene positions

```{r}
## Interpolate CC Diplotype Probabilities to Gene Positions
## This script uses the interpolation code to map marker-based probabilities to gene positions

# Load required libraries
library(IRanges)
library(qtl2convert) # You might need to install this
library(qtl2)        # You might need to install this
library(data.table)
library(dplyr)

# Define interpolation functions
interpolate_one_chr = function(pr1, mkr1, mkr2) {
  
  # Verify that the marker names in pr1 and mkr1 are identical and in the same
  # order.
  stopifnot(dimnames(pr1)[[3]] == names(mkr1))
  
  # Verify that the markers are in bp and not Mbp.
  stopifnot(min(mkr1) > 200)
  stopifnot(min(mkr2) > 200)

  # Convert markers to IRanges to used inter-range functions.
  mkr1 = IRanges(start = mkr1, width = 1, names = names(mkr1))
  mkr2 = IRanges(start = mkr2, width = 1, names = names(mkr2))
  
  # The the proximal and distal markers in mkr1 which bracket the markers
  # in mkr2.
  proximal = follow(mkr2,  mkr1)
  distal   = precede(mkr2, mkr1)
  proximal[is.na(proximal)] = 1
  distal[is.na(distal)]     = length(mkr1)
  
  # Ensure indices are within the range
  proximal[proximal > dim(pr1)[3]] = dim(pr1)[3]
  distal[distal     > dim(pr1)[3]] = dim(pr1)[3]
  
  # Get the proximal and distal probs.
  prox_pr = pr1[,,proximal, drop = FALSE]
  dist_pr = pr1[,,distal,   drop = FALSE]
  
  # interpolate values
  mkr_interp = (start(mkr2) - start(mkr1)[proximal]) / (start(mkr1)[distal] - start(mkr1)[proximal])
  mkr_interp[is.infinite(mkr_interp)] = 1
  mkr_interp = array(rep(mkr_interp, each = nrow(pr1) * ncol(pr1)),
                     c(nrow(pr1), ncol(pr1), length(mkr2)),
                     dimnames = list(rownames(pr1), colnames(pr1), names(mkr2)))

  # Create the new probs array.
  tmp = prox_pr + (mkr_interp) * (dist_pr - prox_pr)
  dimnames(tmp)[[3]] = names(mkr2)
  return(tmp)
}

# Top-level interpolate function.
interpolate_genoprobs = function(probs1, markers1, markers2) {

  # Perform some minimal error checking. 
  if(length(probs1) != length(markers1)) {
    stop('Probs1 and markers1 must be the same length.')
  }

  # If the marker maps are in Mb, convert to bp.
  if(max(unlist(markers1) < 200)) {
    markers1 = lapply(markers1, '*', 1e6)
  }

  if(max(unlist(markers2) < 200)) {
    markers2 = lapply(markers2, '*', 1e6)
  }
  
  all_chr = names(probs1)

  new_probs1        = vector('list', length(probs1))
  names(new_probs1) = names(probs1)

  for(chr in all_chr) {
    new_probs1[[chr]] = interpolate_one_chr(pr1  = probs1[[chr]],
                                           mkr1 = markers1[[chr]],
                                           mkr2 = markers2[[chr]])
  }

  # Set the attributes for the probs object.
  attr(new_probs1, 'crosstype')    = attr(probs1, 'crosstype')
  attr(new_probs1, 'is_x_chr')     = attr(probs1, 'is_x_chr')
  attr(new_probs1, 'alleles')      = attr(probs1, 'alleles')
  attr(new_probs1, 'alleleprobs')  = attr(probs1, 'alleleprobs')
  class(new_probs1)                = class(probs1)
  
  return(new_probs1)
}

# Step 1: Print info about existing data objects
cat("Examining existing data objects...\n")
cat("Dimensions of annotation data:", dim(anno), "\n")
cat("Dimensions of gene data:", dim(genes_with_id), "\n")

# Show chromosome distributions
cat("\nChromosome distribution in anno:\n")
print(table(anno$chr))

cat("\nChromosome distribution in genes_with_id:\n")
print(table(genes_with_id$chromosome_name))

cat("\nStructure of filtered_CC_diplotype_probs:\n")
for (chr in names(filtered_CC_diplotype_probs)) {
  chr_array <- filtered_CC_diplotype_probs[[chr]]
  cat("Chromosome", chr, "dimensions:", paste(dim(chr_array), collapse=" x "), "\n")
  cat("  Samples:", paste(dimnames(chr_array)[[1]], collapse=", "), "\n")
  cat("  Founders:", paste(dimnames(chr_array)[[2]], collapse=", "), "\n")
  cat("  First few markers:", paste(head(dimnames(chr_array)[[3]], 3), collapse=", "), "\n")
}

# Step 2: Format marker data into qtl2 format
cat("\nFormatting marker data...\n")
markers_list <- list()
skipped_chrs <- character(0)

for (chr in names(filtered_CC_diplotype_probs)) {
  # Get the chromosome number without "chr" prefix
  chr_num <- gsub("^chr", "", chr)
  
  # Find markers for this chromosome in anno
  anno_chr <- anno[anno$chr == chr_num, ]
  
  if (nrow(anno_chr) == 0) {
    cat("Warning: No markers found in anno for chromosome", chr, "\n")
    skipped_chrs <- c(skipped_chrs, chr)
    next
  }
  
  # Get marker positions
  marker_pos <- anno_chr$bp_mm10
  names(marker_pos) <- anno_chr$marker
  
  # Get marker names from probability data
  prob_markers <- dimnames(filtered_CC_diplotype_probs[[chr]])[[3]]
  
  # Check for marker name match
  matching_markers <- intersect(names(marker_pos), prob_markers)
  
  if (length(matching_markers) > 0) {
    cat("Found", length(matching_markers), "matching markers for chromosome", chr, "\n")
    
    # Keep only matching markers in the same order as in probability data
    matching_indices <- match(matching_markers, prob_markers)
    matching_markers_ordered <- prob_markers[sort(matching_indices)]
    
    # Create a subset of markers
    marker_pos_subset <- marker_pos[matching_markers_ordered]
    
    # Store the marker positions
    markers_list[[chr]] <- marker_pos_subset
  } else {
    cat("No matching marker names for chromosome", chr, "\n")
    
    # Instead of using synthetic positions, we'll try to map based on position ordering
    # This assumes that both lists are ordered by genomic position
    marker_count <- length(prob_markers)
    
    if (length(marker_pos) >= marker_count) {
      # Sample positions at regular intervals
      indices <- round(seq(1, length(marker_pos), length.out = marker_count))
      sampled_pos <- marker_pos[indices]
      names(sampled_pos) <- prob_markers
      markers_list[[chr]] <- sampled_pos
      cat("Mapped", length(sampled_pos), "markers using position-based assignment\n")
    } else {
      cat("Warning: Not enough markers in anno for chromosome", chr, "\n")
      cat("Skipping chromosome", chr, "due to insufficient marker data\n")
      skipped_chrs <- c(skipped_chrs, chr)
    }
  }
}

# Report on skipped chromosomes
if (length(skipped_chrs) > 0) {
  cat("\nThe following chromosomes were skipped due to insufficient marker data:\n")
  cat(paste(skipped_chrs, collapse=", "), "\n")
}

# Update to remove chromosomes with insufficient data
valid_chrs <- names(markers_list)
cat("\nChromosomes with valid marker data:", paste(valid_chrs, collapse=", "), "\n")

# Filter the probability data to only include chromosomes with sufficient marker data
# Don't use direct subsetting, which can break attributes
new_filtered_probs <- vector('list', length(valid_chrs))
names(new_filtered_probs) <- valid_chrs

for (chr in valid_chrs) {
  new_filtered_probs[[chr]] <- filtered_CC_diplotype_probs[[chr]]
}

# Copy over all attributes
attr(new_filtered_probs, 'crosstype') <- attr(filtered_CC_diplotype_probs, 'crosstype')
attr(new_filtered_probs, 'is_x_chr') <- attr(filtered_CC_diplotype_probs, 'is_x_chr')[valid_chrs]
attr(new_filtered_probs, 'alleles') <- attr(filtered_CC_diplotype_probs, 'alleles')
attr(new_filtered_probs, 'alleleprobs') <- attr(filtered_CC_diplotype_probs, 'alleleprobs')
class(new_filtered_probs) <- class(filtered_CC_diplotype_probs)

# Replace the original object
filtered_CC_diplotype_probs <- new_filtered_probs

# Step 3: Format gene positions into qtl2 format
cat("\nFormatting gene positions...\n")
genes_list <- list()
genes_skipped_chrs <- character(0)

for (chr in names(filtered_CC_diplotype_probs)) {
  # Get the chromosome number without "chr" prefix
  chr_num <- gsub("^chr", "", chr)
  
  # Find genes for this chromosome
  # Try different formats depending on how chromosome names are stored in genes_with_id
  gene_chr_formats <- c(
    chr_num,                # e.g., "1"
    chr,                    # e.g., "chr1"
    toupper(chr_num),       # e.g., "X"
    paste0("chr", toupper(chr_num))  # e.g., "chrX"
  )
  
  genes_found <- FALSE
  
  for (gene_chr in gene_chr_formats) {
    genes_chr <- genes_with_id[genes_with_id$chromosome_name == gene_chr, ]
    
    if (nrow(genes_chr) > 0) {
      cat("Found", nrow(genes_chr), "genes for chromosome", chr, "using format", gene_chr, "\n")
      
      # Use midpoint of gene as position
      gene_pos <- (genes_chr$start_position + genes_chr$end_position) / 2
      names(gene_pos) <- genes_chr$ensembl_gene_id
      
      # Store the genes for this chromosome
      genes_list[[chr]] <- gene_pos
      genes_found <- TRUE
      break
    }
  }
  
  if (!genes_found) {
    cat("Warning: No genes found for chromosome", chr, "\n")
    genes_skipped_chrs <- c(genes_skipped_chrs, chr)
  }
}

# Report on chromosomes with no gene data
if (length(genes_skipped_chrs) > 0) {
  cat("\nThe following chromosomes were skipped due to missing gene data:\n")
  cat(paste(genes_skipped_chrs, collapse=", "), "\n")
}

# Step 4: Check alignment between markers and genes
cat("\nChecking alignment between markers and genes...\n")
common_chrs <- intersect(names(markers_list), names(genes_list))
cat("Common chromosomes for interpolation:", paste(common_chrs, collapse=", "), "\n")

if (length(common_chrs) == 0) {
  stop("No common chromosomes found between markers and genes!")
}

# Filter to only common chromosomes
# Create new lists instead of subsetting the original objects
new_markers_list <- list()
new_genes_list <- list()
probs_list <- list()

for (chr in common_chrs) {
  new_markers_list[[chr]] <- markers_list[[chr]]
  new_genes_list[[chr]] <- genes_list[[chr]]
  probs_list[[chr]] <- filtered_CC_diplotype_probs[[chr]]
}

# Copy over attributes to probs_list
attr(probs_list, 'crosstype') <- attr(filtered_CC_diplotype_probs, 'crosstype')
attr(probs_list, 'is_x_chr') <- attr(filtered_CC_diplotype_probs, 'is_x_chr')[common_chrs]
attr(probs_list, 'alleles') <- attr(filtered_CC_diplotype_probs, 'alleles')
attr(probs_list, 'alleleprobs') <- attr(filtered_CC_diplotype_probs, 'alleleprobs')
class(probs_list) <- class(filtered_CC_diplotype_probs)

# Step 5: Ensure marker alignment for each chromosome
cat("\nChecking marker alignment for each chromosome...\n")
for (chr in common_chrs) {
  prob_markers <- dimnames(probs_list[[chr]])[[3]]
  marker_pos_markers <- names(new_markers_list[[chr]])
  
  if (!identical(prob_markers, marker_pos_markers)) {
    cat("Aligning markers for chromosome", chr, "\n")
    
    # Find the common markers
    common_markers <- intersect(prob_markers, marker_pos_markers)
    
    if (length(common_markers) > 0) {
      # Keep only common markers in both datasets
      prob_indices <- match(common_markers, prob_markers)
      probs_list[[chr]] <- probs_list[[chr]][,,prob_indices, drop=FALSE]
      
      # Keep the same markers in marker_pos
      new_markers_list[[chr]] <- new_markers_list[[chr]][common_markers]
      
      cat("  Aligned", length(common_markers), "markers for chromosome", chr, "\n")
    } else {
      # If no common markers but markers are the same length, we can try to use position ordering
      if (length(prob_markers) == length(marker_pos_markers)) {
        # If same length, just rename the marker positions
        names(new_markers_list[[chr]]) <- prob_markers
        cat("  Renamed markers in position list to match probability markers\n")
      } else {
        # If different lengths and no common markers, this is problematic
        cat("Warning: Cannot establish marker correspondence for chromosome", chr, "\n")
        cat("  Removing chromosome", chr, "from analysis\n")
        
        # Remove this chromosome from our lists
        new_markers_list[[chr]] <- NULL
        new_genes_list[[chr]] <- NULL
        probs_list[[chr]] <- NULL
      }
    }
  }
}

# Update common_chrs after alignment checks
common_chrs <- intersect(names(new_markers_list), names(new_genes_list))
cat("\nFinal chromosomes for interpolation after alignment checks:", paste(common_chrs, collapse=", "), "\n")

if (length(common_chrs) == 0) {
  stop("No common chromosomes left after alignment checks!")
}

# Double-check marker alignment for all remaining chromosomes
for (chr in common_chrs) {
  if (!identical(dimnames(probs_list[[chr]])[[3]], names(new_markers_list[[chr]]))) {
    stop("Failed to align markers for chromosome ", chr)
  }
}

# Step 6: Apply interpolation
cat("\nPerforming interpolation...\n")
interpolated_probs <- interpolate_genoprobs(
  probs1 = probs_list,
  markers1 = new_markers_list,
  markers2 = new_genes_list
)

# Step 7: Convert to a more convenient format for downstream analysis
cat("\nConverting interpolated probabilities to data frame...\n")
interpolated_df <- data.frame()

for (chr_name in names(interpolated_probs)) {
  chr_array <- interpolated_probs[[chr_name]]
  
  # Extract dimensions
  samples <- dimnames(chr_array)[[1]]
  founders <- dimnames(chr_array)[[2]]
  genes <- dimnames(chr_array)[[3]]
  
  cat("Processing chromosome", chr_name, "with", length(genes), "genes\n")
  
  # Create data frame more efficiently with pre-allocation
  chr_rows <- length(samples) * length(founders) * length(genes)
  chr_df <- data.frame(
    sample = character(chr_rows),
    founder = character(chr_rows),
    gene = character(chr_rows),
    chromosome = character(chr_rows),
    probability = numeric(chr_rows),
    stringsAsFactors = FALSE
  )
  
  row_idx <- 1
  for (sample_idx in 1:length(samples)) {
    sample <- samples[sample_idx]
    for (founder_idx in 1:length(founders)) {
      founder <- founders[founder_idx]
      
      # Process batches of genes for efficiency
      batch_size <- 1000
      for (start_idx in seq(1, length(genes), by = batch_size)) {
        end_idx <- min(start_idx + batch_size - 1, length(genes))
        batch_genes <- genes[start_idx:end_idx]
        batch_size_actual <- length(batch_genes)
        
        # Extract batch probabilities
        batch_probs <- chr_array[sample_idx, founder_idx, batch_genes]
        
        # Assign to pre-allocated dataframe
        batch_indices <- row_idx:(row_idx + batch_size_actual - 1)
        chr_df$sample[batch_indices] <- sample
        chr_df$founder[batch_indices] <- founder
        chr_df$gene[batch_indices] <- batch_genes
        chr_df$chromosome[batch_indices] <- sub("chr", "", chr_name)
        chr_df$probability[batch_indices] <- batch_probs
        
        row_idx <- row_idx + batch_size_actual
      }
    }
  }
  
  # Append to main dataframe
  interpolated_df <- rbind(interpolated_df, chr_df[1:(row_idx-1),])
}

# Step 8: Save the interpolated probabilities with timestamp
cat("\nSaving results...\n")
save_rds_file <- paste0("Results/Interpolate/CC_gene_interpolated_probs.rds")

# Save both the array and dataframe formats
saveRDS(interpolated_probs, save_rds_file)
save_csv_file <- paste0("Results/Interpolate/CC_gene_interpolated_probs.csv")
data.table::fwrite(interpolated_df, save_csv_file)

cat("Saved interpolated probabilities to", save_rds_file, "\n")
cat("Saved interpolated probabilities as dataframe to", save_csv_file, "\n")

# Final report
cat("\nSummary of interpolation:\n")
cat("- Started with", length(names(filtered_CC_diplotype_probs)), "chromosomes in probability data\n")
cat("- Found marker data for", length(names(markers_list)), "chromosomes\n")
cat("- Found gene data for", length(names(genes_list)), "chromosomes\n")
cat("- Successfully interpolated", length(names(interpolated_probs)), "chromosomes\n")

```

## 5. Compare GBRS-derived CC diplotypes vs "CC_gene_interpolated_probs" – Cosine similarity

```{r}
################################################################################
# CC Parental Baseline Cosine Similarity Analysis
# Analyzes CC parental samples for baseline metrics
################################################################################

library(dplyr)
library(ggplot2)
library(tidyr)

# Fixed cosine similarity function with proper error handling
cos_sim = function(a, b) {
  v = rep(1L, ncol(a))
  numerator <- (a * b) %*% v
  denominator <- sqrt(a^2 %*% v * b^2 %*% v)
  
  # Handle division by zero
  result <- numerator / denominator
  
  # Replace NaN values with 0 (no similarity)
  result[is.nan(result)] <- 0
  
  return(mean(result, na.rm = TRUE))
}

# Enhanced cosine similarity that also returns similarity per gene
cos_sim_per_gene = function(a, b) {
  v = rep(1L, ncol(a))
  numerator <- (a * b) %*% v
  denominator <- sqrt(a^2 %*% v * b^2 %*% v)
  
  # Handle division by zero
  gene_sims <- numerator / denominator
  
  # Replace NaN values with 0 (no similarity)
  gene_sims[is.nan(gene_sims)] <- 0
  
  return(list(
    overall_sim = mean(gene_sims, na.rm = TRUE),
    per_gene_sim = gene_sims
  ))
}

# Convert GBRS read counts to allele proportions matrix
gbrs_to_proportions <- function(gbrs_counts) {
  # Initialize the proportions matrix
  prop_matrix <- matrix(0, nrow = nrow(gbrs_counts), ncol = 8)
  rownames(prop_matrix) <- gbrs_counts$locus
  colnames(prop_matrix) <- LETTERS[1:8]
  
  # Fill in proportions
  for (i in 1:nrow(gbrs_counts)) {
    if (!is.na(gbrs_counts$total[i]) && gbrs_counts$total[i] > 0) {
      prop_matrix[i, ] <- as.numeric(gbrs_counts[i, LETTERS[1:8]]) / gbrs_counts$total[i]
    }
  }
  
  return(prop_matrix)
}

# Updated extract_strain_probs function for the specific CC_gene_interpolated_probs structure
extract_strain_probs <- function(interpolated_probs, strain_name, gene_ids) {
  # Create matrix for this strain
  strain_matrix <- matrix(0, nrow = length(gene_ids), ncol = 8)
  rownames(strain_matrix) <- gene_ids
  colnames(strain_matrix) <- LETTERS[1:8]
  
  # Find the strain's index in each chromosome's data
  for (chr in names(interpolated_probs)) {
    # Get strain index (row index) - assuming CC_gene_interpolated_probs structure
    # where first dimension is strain, with strain names in dimnames[[1]]
    strain_idx <- which(dimnames(interpolated_probs[[chr]])[[1]] == strain_name)
    
    if (length(strain_idx) > 0) {
      # Get genes for this chromosome
      chr_genes <- intersect(gene_ids, dimnames(interpolated_probs[[chr]])[[3]])
      
      # Extract probabilities for this strain and these genes
      for (gene in chr_genes) {
        if (gene %in% dimnames(interpolated_probs[[chr]])[[3]]) {
          gene_idx <- match(gene, dimnames(interpolated_probs[[chr]])[[3]])
          # Extract the probability array for this strain and gene
          strain_matrix[gene, ] <- interpolated_probs[[chr]][strain_idx[1], , gene_idx]
        }
      }
    }
  }
  
  return(strain_matrix)
}

# Analyze cosine similarity for a single CC parental sample
analyze_sample_cosine <- function(sample_id, gbrs_base_path, interpolated_probs, strain_id) {
  # Construct file path for CC parental sample
  sample_file <- file.path(gbrs_base_path, sample_id, 
                          paste0(sample_id, "_gbrs.quantified.diploid.genes.expected_read_counts"))
  
  # Check if file exists
  if (!file.exists(sample_file)) {
    cat("File not found for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Load GBRS alignment counts
  gbrs_counts <- tryCatch({
    read.table(sample_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  }, error = function(e) {
    cat("Error reading file for sample", sample_id, ":", e$message, "\n")
    return(NULL)
  })
  
  if (is.null(gbrs_counts) || nrow(gbrs_counts) == 0) {
    cat("Empty file for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Check required columns
  required_cols <- c("locus", "total", LETTERS[1:8])
  if (!all(required_cols %in% colnames(gbrs_counts))) {
    cat("Missing required columns for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Convert GBRS data to proportions matrix
  sample_matrix <- gbrs_to_proportions(gbrs_counts)
  
  # Get gene IDs to analyze (those in the GBRS file with non-zero totals)
  valid_genes <- gbrs_counts$locus[!is.na(gbrs_counts$total) & gbrs_counts$total > 0]
  
  # Extract strain matrix for the parental strain
  strain_matrix <- extract_strain_probs(interpolated_probs, strain_id, valid_genes)
  
  # Ensure matrices have the same rows (genes)
  common_genes <- intersect(rownames(sample_matrix), rownames(strain_matrix))
  
  if (length(common_genes) == 0) {
    cat("No common genes found for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Subset matrices to common genes
  sample_matrix <- sample_matrix[common_genes, ]
  strain_matrix <- strain_matrix[common_genes, ]
  
  # Remove rows that are all zeros in any matrix
  non_zero_rows <- rowSums(sample_matrix) > 0 & rowSums(strain_matrix) > 0
  
  if (sum(non_zero_rows) == 0) {
    cat("ERROR: No valid non-zero rows found for comparison\n")
    return(NULL)
  }
  
  sample_matrix <- sample_matrix[non_zero_rows, ]
  strain_matrix <- strain_matrix[non_zero_rows, ]
  
  # Calculate similarity
  sim_to_strain <- cos_sim(sample_matrix, strain_matrix)
  
  # Calculate per-gene similarities for detailed analysis
  detailed_sim_strain <- cos_sim_per_gene(sample_matrix, strain_matrix)
  
  # Calculate similarity thresholds for the cosine similarity
  gene_similarities <- data.frame(
    gene = rownames(sample_matrix),
    sim_strain = detailed_sim_strain$per_gene_sim,
    stringsAsFactors = FALSE
  )
  
  # Map genes to chromosomes using the interpolated_probs structure
  chromosomes <- rep("unknown", length(gene_similarities$gene))
  names(chromosomes) <- gene_similarities$gene
  
  # For each chromosome in the interpolated_probs list
  for (chr in names(interpolated_probs)) {
    # Get genes in this chromosome from third dimension of array
    chr_genes <- dimnames(interpolated_probs[[chr]])[[3]]
    
    # Find matching genes between our gene set and this chromosome
    matching_genes <- intersect(gene_similarities$gene, chr_genes)
    
    # Set chromosome for these genes
    if (length(matching_genes) > 0) {
      chromosomes[matching_genes] <- paste0("chr", chr)
    }
  }
  
  # Add chromosome information to the gene_similarities dataframe
  gene_similarities$chromosome <- chromosomes[gene_similarities$gene]
  
  # Calculate per-chromosome similarity
  chr_similarities <- gene_similarities %>%
    dplyr::group_by(chromosome) %>%
    dplyr::summarize(
      sim_strain = mean(sim_strain, na.rm = TRUE),
      gene_count = dplyr::n()
    )
  
  # Create final result object
  result <- list(
    sample_id = sample_id,
    strain_id = strain_id,
    n_genes = nrow(sample_matrix),
    sim_to_strain = sim_to_strain,
    
    # Detailed results for further analysis
    gene_similarities = gene_similarities,
    chr_similarities = chr_similarities
  )
  
  return(result)
}

# Modified function to handle alpha-numeric sample IDs
analyze_all_parental_samples <- function(sample_list, gbrs_base_path, metadata_subset, 
                                     interpolated_probs, parental_strains) {
  results <- list()
  
  # Initialize results list for each CC line
  for (strain in unique(names(parental_strains))) {
    results[[strain]] <- list()
  }
  
  for (sample_id in sample_list) {
    # Find this sample in the metadata using exact matching of ID as character
    # This avoids numeric conversion issues with alpha-numeric IDs
    sample_row <- which(as.character(metadata_subset$Animal.ID) == sample_id)
    
    if (length(sample_row) == 0) {
      cat("Skipping sample", sample_id, "- not found in metadata\n")
      next
    }
    
    # Get the cross type from metadata
    cross_type <- metadata_subset$Cross[sample_row[1]]
    
    # Skip if cross type not in parental_strains
    if (!(cross_type %in% names(parental_strains))) {
      cat("Skipping sample", sample_id, "- cross type", cross_type, "not defined in parental_strains\n")
      next
    }
    
    # Get strain ID for this cross
    strain_id <- parental_strains[[cross_type]]
    
    cat("Processing sample", sample_id, "(Cross:", cross_type, ", Strain:", strain_id, ")\n")
    
    # Analyze this sample against its parental strain
    sample_result <- analyze_sample_cosine(sample_id, gbrs_base_path, 
                                          interpolated_probs, strain_id)
    
    if (!is.null(sample_result)) {
      results[[cross_type]][[sample_id]] <- sample_result
      cat("Analyzed", sample_result$n_genes, "genes for sample", sample_id, "\n")
      cat("Cosine similarity to", strain_id, ":", round(sample_result$sim_to_strain, 3), "\n")
    }
  }
  
  return(results)
}

# Summarize cosine similarity results for parental samples
summarize_parental_results <- function(cosine_results, parental_strains) {
  summary <- list()
  
  for (strain_cross in names(cosine_results)) {
    # Skip if no samples for this strain
    if (length(cosine_results[[strain_cross]]) == 0) {
      next
    }
    
    cat("\n=======================================\n")
    cat("Parental Strain:", strain_cross, "\n")
    
    # Get strain ID for this cross
    strain_id <- parental_strains[[strain_cross]]
    
    cat("Strain ID:", strain_id, "\n")
    
    # Collect statistics
    sample_count <- length(cosine_results[[strain_cross]])
    cat("Number of samples:", sample_count, "\n")
    
    # Aggregate results
    sim_to_strain <- sapply(cosine_results[[strain_cross]], function(x) x$sim_to_strain)
    genes_analyzed <- sapply(cosine_results[[strain_cross]], function(x) x$n_genes)
    
    cat("Total genes analyzed:", sum(genes_analyzed), "\n")
    cat("Average genes per sample:", round(mean(genes_analyzed)), "\n\n")
    
    cat("=== COSINE SIMILARITY METRICS ===\n")
    cat("Average similarity to", strain_id, ":", round(mean(sim_to_strain), 3), "\n")
    cat("Min similarity:", round(min(sim_to_strain), 3), "\n")
    cat("Max similarity:", round(max(sim_to_strain), 3), "\n")
    cat("Standard deviation:", round(sd(sim_to_strain), 3), "\n\n")
    
    # Create data frame for sample-level results
    sample_df <- do.call(rbind, lapply(names(cosine_results[[strain_cross]]), function(sample) {
      result <- cosine_results[[strain_cross]][[sample]]
      df <- data.frame(
        Sample = sample,
        Genes = result$n_genes,
        Sim_To_Strain = round(result$sim_to_strain, 3),
        stringsAsFactors = FALSE
      )
      return(df)
    }))
    
    # Aggregate chromosome-level results
    chr_df <- do.call(rbind, lapply(names(cosine_results[[strain_cross]]), function(sample) {
      result <- cosine_results[[strain_cross]][[sample]]
      if (!is.null(result$chr_similarities)) {
        df <- result$chr_similarities
        df$Sample <- sample
        return(df)
      } else {
        return(NULL)
      }
    }))
    
    # Store summary results
    summary[[strain_cross]] <- list(
      strain_cross = strain_cross,
      strain_id = strain_id,
      sample_count = sample_count,
      total_genes = sum(genes_analyzed),
      avg_genes_per_sample = mean(genes_analyzed),
      
      # Cosine similarity metrics
      avg_sim_to_strain = mean(sim_to_strain),
      min_sim_to_strain = min(sim_to_strain),
      max_sim_to_strain = max(sim_to_strain),
      sd_sim_to_strain = sd(sim_to_strain),
      
      # Sample-level results
      sample_results = sample_df,
      
      # Chromosome-level results
      chr_results = chr_df
    )
    
    # Print sample results
    cat("\n=== SAMPLE-LEVEL RESULTS ===\n")
    print(sample_df)
    
    # Print chromosome results summary
    cat("\n=== CHROMOSOME-LEVEL RESULTS ===\n")
    chr_summary <- chr_df %>%
      dplyr::group_by(chromosome) %>%
      dplyr::summarize(
        avg_sim = mean(sim_strain, na.rm = TRUE),
        sd_sim = sd(sim_strain, na.rm = TRUE),
        avg_genes = mean(gene_count, na.rm = TRUE)
      )
    print(chr_summary)
  }
  
  return(summary)
}

# Modified visualization function to include chromosome-level visualization
create_parental_visualizations <- function(summary, output_base_dir = "Results/CC") {
  viz_dirs <- list()
  
  for (strain_cross in names(summary)) {
    strain_data <- summary[[strain_cross]]
    strain_id <- strain_data$strain_id
    
    # Extract the CC number from the strain string for cleaner labels
    strain_label <- sub("/.*", "", strain_id)
    
    # Create strain-specific directory structure: Results/CC/[strain]/Cosine
    strain_dir <- file.path(output_base_dir, strain_cross, "Cosine")
    if (!dir.exists(strain_dir)) {
      dir.create(strain_dir, recursive = TRUE)
    }
    
    # Store the visualization directory for this strain
    viz_dirs[[strain_cross]] <- strain_dir
    
    # Create visualizations only if we have sample results
    if (nrow(strain_data$sample_results) > 0) {
      # 1. Sample-by-sample bar plot
      sample_data <- strain_data$sample_results %>%
        dplyr::mutate(
          Sample = factor(Sample, levels = Sample[order(Sim_To_Strain, decreasing = TRUE)])
        )
      
      sample_plot <- ggplot(sample_data, aes(x = Sample, y = Sim_To_Strain)) +
        geom_bar(stat = "identity", fill = "blue", alpha = 0.7) +
        theme_minimal() +
        labs(
          title = paste("Cosine Similarity by Sample -", strain_cross),
          y = paste("Similarity to", strain_label),
          x = "Sample ID"
        ) +
        geom_hline(yintercept = strain_data$avg_sim_to_strain, 
                  color = "red", linetype = "dashed") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
      
      # Save the sample plot
      sample_plot_file <- file.path(strain_dir, paste0(strain_cross, "_sample_similarities.pdf"))
      ggsave(sample_plot_file, sample_plot, width = 12, height = 6)
      
      # 2. Chromosome-level similarity plot
      if (!is.null(strain_data$chr_results) && nrow(strain_data$chr_results) > 0) {
        # Aggregate by chromosome
        chr_summary <- strain_data$chr_results %>%
          dplyr::group_by(chromosome) %>%
          dplyr::summarize(
            avg_sim = mean(sim_strain, na.rm = TRUE),
            sd_sim = sd(sim_strain, na.rm = TRUE),
            count = dplyr::n()
          )
        
        # Sort chromosomes to ensure proper ordering (chr1, chr2, ..., chr19, chrX)
        chr_summary$chromosome <- factor(chr_summary$chromosome, 
                                        levels = c(paste0("chr", 1:19), "chrX"))
        
        chr_summary <- chr_summary %>% arrange(chromosome)
        
        # Create chromosome plot
        chr_plot <- ggplot(chr_summary, aes(x = chromosome, y = avg_sim)) +
          geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
          geom_errorbar(aes(ymin = avg_sim - sd_sim, ymax = avg_sim + sd_sim), width = 0.2) +
          theme_minimal() +
          labs(
            title = paste("Cosine Similarity by Chromosome -", strain_cross),
            y = paste("Similarity to", strain_label),
            x = "Chromosome"
          ) +
          geom_hline(yintercept = strain_data$avg_sim_to_strain, 
                    color = "red", linetype = "dashed") +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))
        
        # Save the chromosome plot
        chr_plot_file <- file.path(strain_dir, paste0(strain_cross, "_chromosome_similarities.pdf"))
        ggsave(chr_plot_file, chr_plot, width = 10, height = 6)
      }
    }
    
    cat("Visualizations for strain", strain_cross, "created in directory:", strain_dir, "\n")
  }
  
  # Create a comparison plot if we have multiple strains
  if (length(summary) > 1) {
    # Prepare comparison data for chromosome-level comparison across strains
    chr_comparison_data <- data.frame(
      Strain = character(),
      Chromosome = character(),
      Similarity = numeric(),
      SD = numeric(),
      stringsAsFactors = FALSE
    )
    
    for (strain_cross in names(summary)) {
      strain_data <- summary[[strain_cross]]
      if (!is.null(strain_data$chr_results) && nrow(strain_data$chr_results) > 0) {
        # Aggregate chromosome data for this strain
        chr_summary <- strain_data$chr_results %>%
          dplyr::group_by(chromosome) %>%
          dplyr::summarize(
            avg_sim = mean(sim_strain, na.rm = TRUE),
            sd_sim = sd(sim_strain, na.rm = TRUE)
          )
        
        temp_df <- data.frame(
          Strain = rep(strain_cross, nrow(chr_summary)),
          Chromosome = chr_summary$chromosome,
          Similarity = chr_summary$avg_sim,
          SD = chr_summary$sd_sim,
          stringsAsFactors = FALSE
        )
        
        chr_comparison_data <- rbind(chr_comparison_data, temp_df)
      }
    }
    
    # Create chromosome comparison plot if we have data
    if (nrow(chr_comparison_data) > 0) {
      # Ensure proper chromosome ordering
      chr_comparison_data$Chromosome <- factor(chr_comparison_data$Chromosome, 
                                            levels = c(paste0("chr", 1:19), "chrX"))
      
      # Create the comparison plot
      chr_compare_plot <- ggplot(chr_comparison_data, 
                              aes(x = Chromosome, y = Similarity, fill = Strain)) +
        geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
        geom_errorbar(aes(ymin = Similarity - SD, ymax = Similarity + SD), 
                     position = position_dodge(width = 0.9), width = 0.2) +
        theme_minimal() +
        labs(
          title = "Chromosome Similarity Comparison Across Strains",
          y = "Cosine Similarity",
          x = "Chromosome"
        ) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
      # Save the chromosome comparison plot
      chr_compare_file <- file.path(output_base_dir, "chromosome_strain_comparison.pdf")
      ggsave(chr_compare_file, chr_compare_plot, width = 12, height = 6)
      
      cat("Chromosome comparison plot created at:", chr_compare_file, "\n")
    }
  }
  
  return(viz_dirs)
}

#################################################################################
# EXAMPLE USAGE SCRIPT USING PRELOADED VARIABLES
#################################################################################

# We assume the following variables are preloaded:
# - metadata_subset: Contains sample metadata (Animal.ID, Cross, etc.)
# - CC_gene_interpolated_probs: Contains interpolated probabilities for CC strains

# Define sample list from metadata - ensure they're character strings
sample_list <- as.character(metadata_subset$Animal.ID)

# Define base path to GBRS data
gbrs_base_path <- "CC_quantify_ASE"

# Define output directory base
output_base_dir <- "Results/CC"

# Define parental strains for each CC cross type
parental_strains <- list(
  "CC001xCC001" = "CC001/Unc",
  "CC019xCC019" = "CC019/TauUnc",
  "CC040xCC040" = "CC040/TauUnc"
)

# Run the analysis
results <- analyze_all_parental_samples(
  sample_list = sample_list,
  gbrs_base_path = gbrs_base_path,
  metadata_subset = metadata_subset,
  interpolated_probs = CC_gene_interpolated_probs,
  parental_strains = parental_strains
)

# Summarize results
summary <- summarize_parental_results(results, parental_strains)

# Create visualizations - saves to Results/CC/[strain]/Cosine
viz_dirs <- create_parental_visualizations(summary, output_base_dir)

# Generate summary reports for each strain
for (strain_cross in names(summary)) {
  # Create strain-specific directory for output
  strain_cosine_dir <- file.path(output_base_dir, strain_cross, "Cosine")
  if (!dir.exists(strain_cosine_dir)) {
    dir.create(strain_cosine_dir, recursive = TRUE)
  }
  
  # Generate report for this strain
  strain_data <- summary[[strain_cross]]
  report_file <- file.path(strain_cosine_dir, paste0(strain_cross, "_cosine_analysis_report.txt"))
  
  sink(report_file)
  cat("CC Parental Baseline Cosine Similarity Analysis for", strain_cross, "\n")
  cat(paste(rep("=", 50), collapse = ""), "\n\n")
  cat("Analysis Date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
  
  # Write strain-specific information
  cat("\nCC Strain Cross:", strain_cross, "\n")
  cat("Reference Strain ID:", strain_data$strain_id, "\n")
  cat("Number of Samples:", strain_data$sample_count, "\n")
  cat("Total Genes Analyzed:", strain_data$total_genes, "\n")
  cat("Average Genes per Sample:", round(strain_data$avg_genes_per_sample), "\n\n")
  
  cat("COSINE SIMILARITY METRICS:\n")
  cat("Average similarity to reference strain:", round(strain_data$avg_sim_to_strain, 3), "\n")
  cat("Minimum similarity:", round(strain_data$min_sim_to_strain, 3), "\n")
  cat("Maximum similarity:", round(strain_data$max_sim_to_strain, 3), "\n")
  cat("Standard deviation:", round(strain_data$sd_sim_to_strain, 3), "\n\n")
  
  # Print sample-level results
  cat("SAMPLE-LEVEL RESULTS:\n")
  print(strain_data$sample_results)
  
  # Add chromosome-level results to the report
  if (!is.null(strain_data$chr_results) && nrow(strain_data$chr_results) > 0) {
    cat("\nCHROMOSOME-LEVEL RESULTS:\n")
    chr_summary <- strain_data$chr_results %>%
      group_by(chromosome) %>%
      summarize(
        avg_sim = mean(sim_strain, na.rm = TRUE),
        sd_sim = sd(sim_strain, na.rm = TRUE),
        avg_genes = mean(gene_count, na.rm = TRUE)
      )
    print(chr_summary)
  }
  
  sink()
  
  cat("Report for strain", strain_cross, "generated at", report_file, "\n")
  
  # Also save the summary as an RDS file
  summary_file <- file.path(strain_cosine_dir, paste0(strain_cross, "_cosine_summary.rds"))
  saveRDS(strain_data, summary_file)
  
  # Save the raw results for this strain as well
  if (strain_cross %in% names(results)) {
    raw_results_file <- file.path(strain_cosine_dir, paste0(strain_cross, "_cosine_raw_results.rds"))
    saveRDS(results[[strain_cross]], raw_results_file)
  }
}

# Create overall summary file
overall_summary_file <- file.path(output_base_dir, "cc_parental_baseline_summary.txt")
sink(overall_summary_file)
cat("CC Parental Baseline Cosine Similarity Summary\n")
cat("============================================\n\n")
cat("Analysis Date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n\n")

# Create a comparison table
comparison_table <- data.frame(
  CC_Strain = character(),
  Avg_Similarity = numeric(),
  Min_Similarity = numeric(),
  Max_Similarity = numeric(),
  SD_Similarity = numeric(),
  Sample_Count = integer(),
  stringsAsFactors = FALSE
)

for (strain_cross in names(summary)) {
  strain_data <- summary[[strain_cross]]
  strain_label <- sub("/.*", "", strain_data$strain_id)
  
  comparison_table <- rbind(comparison_table, data.frame(
    CC_Strain = strain_cross,
    Avg_Similarity = round(strain_data$avg_sim_to_strain, 3),
    Min_Similarity = round(strain_data$min_sim_to_strain, 3),
    Max_Similarity = round(strain_data$max_sim_to_strain, 3),
    SD_Similarity = round(strain_data$sd_sim_to_strain, 3),
    Sample_Count = strain_data$sample_count,
    stringsAsFactors = FALSE
  ))
}

cat("COMPARISON OF PARENTAL STRAINS:\n")
print(comparison_table)
sink()

cat("Overall summary generated at", overall_summary_file, "\n")

# Save overall results as RDS for future reference


# CC001: 0.947
# CC019: 0.909
# CC040: 0.934
```

## 6. Generate CCRIX combined parental diplotype matrices

```{r}
# Updated function to generate sex-specific combined matrices for RIX combinations
generate_rix_combined_matrices_by_sex <- function(interpolated_probs, output_dir = "Results/Combined_Matrices_By_Sex") {
  # Make sure output directory exists
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Define the RIX combinations
  rix_combinations <- list(
    "RIX1" = c("CC001/Unc", "CC019/TauUnc"),
    "RIX2" = c("CC019/TauUnc", "CC040/TauUnc"),
    "RIX3" = c("CC001/Unc", "CC040/TauUnc")
  )
  
  # Function to create combined matrix for a chromosome, taking sex into account
  create_combined_chromosome <- function(chromosome_data, parent1_idx, parent2_idx, is_female, is_x_chr) {
    # Get the number of genes in this chromosome
    n_genes <- dim(chromosome_data)[3]
    
    # Create a matrix to store the combined probabilities
    combined_matrix <- array(0, dim = c(1, dim(chromosome_data)[2], n_genes))
    
    # Name the dimensions appropriately
    dimnames(combined_matrix) <- list(
      "Combined",
      dimnames(chromosome_data)[[2]], 
      dimnames(chromosome_data)[[3]]
    )
    
    # For each gene, combine parent probabilities based on sex and chromosome
    for (gene_idx in 1:n_genes) {
      parent1_probs <- chromosome_data[parent1_idx, , gene_idx]
      parent2_probs <- chromosome_data[parent2_idx, , gene_idx]
      
      # For X chromosome in males, only use maternal X
      if (!is_female && is_x_chr) {
        # For males, X chromosome is only from the mother (parent1 in forward cross)
        combined_matrix[1, , gene_idx] <- parent1_probs
      } else {
        # For autosomes (both sexes) and X in females, average both parents
        combined_matrix[1, , gene_idx] <- (parent1_probs + parent2_probs) / 2
      }
    }
    
    return(combined_matrix)
  }
  
  # Process each RIX combination for both forward and reverse crosses (to account for sex differences)
  for (rix_name in names(rix_combinations)) {
    # Forward cross (parent1 = female, parent2 = male)
    cat("Processing", rix_name, "Forward Cross...\n")
    
    parent1 <- rix_combinations[[rix_name]][1]  # mother
    parent2 <- rix_combinations[[rix_name]][2]  # father
    
    # Process both sexes for forward cross
    for (sex in c("Female", "Male")) {
      is_female <- sex == "Female"
      cross_direction <- "Forward"
      
      cat("  Creating", sex, "matrix for", cross_direction, "cross:", parent1, "(♀) ×", parent2, "(♂)\n")
      
      # Create a new interpolated_probs structure for the combined data
      combined_probs <- list()
      
      # Process each chromosome
      for (chr in names(interpolated_probs)) {
        cat("    Processing chromosome", chr, "\n")
        
        # Determine if this is X chromosome
        is_x_chr <- chr == "X" || chr == "chrX"
        
        # Find parent indices in this chromosome
        parent1_idx <- which(rownames(interpolated_probs[[chr]]) == parent1)
        parent2_idx <- which(rownames(interpolated_probs[[chr]]) == parent2)
        
        if (length(parent1_idx) == 0 || length(parent2_idx) == 0) {
          cat("    Warning: Could not find both parents in chromosome", chr, "\n")
          next
        }
        
        # Create combined matrix for this chromosome, accounting for sex and X chromosome
        combined_probs[[chr]] <- create_combined_chromosome(
          interpolated_probs[[chr]], parent1_idx, parent2_idx, is_female, is_x_chr
        )
      }
      
      # Save the combined matrix
      output_file <- file.path(output_dir, paste0(rix_name, "_", cross_direction, "_", sex, "_combined_matrix.rds"))
      saveRDS(combined_probs, output_file)
      cat("  Saved combined matrix to", output_file, "\n")
    }
    
    # Reverse cross (parent2 = female, parent1 = male)
    cat("Processing", rix_name, "Reverse Cross...\n")
    
    parent1 <- rix_combinations[[rix_name]][2]  # mother in reverse cross
    parent2 <- rix_combinations[[rix_name]][1]  # father in reverse cross
    
    # Process both sexes for reverse cross
    for (sex in c("Female", "Male")) {
      is_female <- sex == "Female"
      cross_direction <- "Reverse"
      
      cat("  Creating", sex, "matrix for", cross_direction, "cross:", parent1, "(♀) ×", parent2, "(♂)\n")
      
      # Create a new interpolated_probs structure for the combined data
      combined_probs <- list()
      
      # Process each chromosome
      for (chr in names(interpolated_probs)) {
        cat("    Processing chromosome", chr, "\n")
        
        # Determine if this is X chromosome
        is_x_chr <- chr == "X" || chr == "chrX"
        
        # Find parent indices in this chromosome
        parent1_idx <- which(rownames(interpolated_probs[[chr]]) == parent1)
        parent2_idx <- which(rownames(interpolated_probs[[chr]]) == parent2)
        
        if (length(parent1_idx) == 0 || length(parent2_idx) == 0) {
          cat("    Warning: Could not find both parents in chromosome", chr, "\n")
          next
        }
        
        # Create combined matrix for this chromosome, accounting for sex and X chromosome
        combined_probs[[chr]] <- create_combined_chromosome(
          interpolated_probs[[chr]], parent1_idx, parent2_idx, is_female, is_x_chr
        )
      }
      
      # Save the combined matrix
      output_file <- file.path(output_dir, paste0(rix_name, "_", cross_direction, "_", sex, "_combined_matrix.rds"))
      saveRDS(combined_probs, output_file)
      cat("  Saved combined matrix to", output_file, "\n")
    }
  }
  
  cat("All RIX combined matrices generated successfully.\n")
  return(output_dir)
}

# Example usage:

```

## 7. Compare GBRS-derived CCRIX diplotypes vs "Combined matrices by Sex" from CC_gene_interpolated_probs – Cosine similarity

```{r}
# RIX1 01x19 
# CC_qunatify_ASE = 0.688
# DO.G10 = 0.890
# DO.G9 = 0.890 
# DO.G8 = 0.890
# DO.G0 = 0.891 !! 
# DO.G100 = 0.889 

# RIX2 19x40
# DO.G10 = 0.889
# DO.G0 = 0.889 
# DO.G100 = 0.887 

# RIX3 01x40
# DO.G10 = 0.906 
# DO.G0 = 0.907 ! 
# DO.G100 = 0.904 

# Best values
# 01x19: 0.891
# 19x40: 0.889
# 01x40: 0.907

# It appears these DO.trabsprob files truly dont have a big influence on the diplotype calls -- appears G0 is the best option (by a margin of 0.001)
```

```{r}
################################################################################
# CCRIX Cosine Similarity Analysis - Using Combined Parental Matrices
# Simplified version - Removed sex comparisons, added single cross capability
################################################################################

library(dplyr)
library(ggplot2)
library(tidyr)

# ============================================================================
# SPECIFY YOUR INPUT FILES AND PARAMETERS HERE
# ============================================================================

# Input file paths - EDIT THESE PATHS to point to your data
metadata_file <- metadata_subset

ccrix_base_path <- "quantify_ASE_G0" 

interpolated_probs_file <- CC_gene_interpolated_probs 

combined_matrices_dir <- "Results/Combined_Matrices_By_Sex"  

# Output settings
output_base_dir <- "Results/CCRIX_G0"

# Optional: Specify a single cross to analyze (e.g., "CC001xCC019")
# Set to NULL to analyze all crosses in the metadata
target_cross <- NULL

# ============================================================================
# HELPER FUNCTIONS
# ============================================================================

# Fixed cosine similarity function with proper error handling
cos_sim = function(a, b) {
  v = rep(1L, ncol(a))
  numerator <- (a * b) %*% v
  denominator <- sqrt(a^2 %*% v * b^2 %*% v)
  
  # Handle division by zero
  result <- numerator / denominator
  
  # Replace NaN values with 0 (no similarity)
  result[is.nan(result)] <- 0
  
  return(mean(result, na.rm = TRUE))
}

# Enhanced cosine similarity that also returns similarity per gene
cos_sim_per_gene = function(a, b) {
  v = rep(1L, ncol(a))
  numerator <- (a * b) %*% v
  denominator <- sqrt(a^2 %*% v * b^2 %*% v)
  
  # Handle division by zero
  gene_sims <- numerator / denominator
  
  # Replace NaN values with 0 (no similarity)
  gene_sims[is.nan(gene_sims)] <- 0
  
  return(list(
    overall_sim = mean(gene_sims, na.rm = TRUE),
    per_gene_sim = gene_sims
  ))
}

# Convert CCRIX read counts to allele proportions matrix
ccrix_to_proportions <- function(ccrix_counts) {
  # Initialize the proportions matrix
  prop_matrix <- matrix(0, nrow = nrow(ccrix_counts), ncol = 8)
  rownames(prop_matrix) <- ccrix_counts$locus
  colnames(prop_matrix) <- LETTERS[1:8]
  
  # Fill in proportions
  for (i in 1:nrow(ccrix_counts)) {
    if (!is.na(ccrix_counts$total[i]) && ccrix_counts$total[i] > 0) {
      prop_matrix[i, ] <- as.numeric(ccrix_counts[i, LETTERS[1:8]]) / ccrix_counts$total[i]
    }
  }
  
  return(prop_matrix)
}

# Extract probability matrices for a strain across all chromosomes
extract_strain_probs <- function(interpolated_probs, strain_name, gene_ids) {
  # Create matrix for this strain
  strain_matrix <- matrix(0, nrow = length(gene_ids), ncol = 8)
  rownames(strain_matrix) <- gene_ids
  
  # Find the strain's index in each chromosome's data
  for (chr in names(interpolated_probs)) {
    # Try exact match
    strain_idx <- which(dimnames(interpolated_probs[[chr]])[[1]] == strain_name)
    
    if (length(strain_idx) > 0) {
      # Get genes for this chromosome
      chr_genes <- intersect(gene_ids, dimnames(interpolated_probs[[chr]])[[3]])
      
      # Extract probabilities for this strain and these genes
      for (gene in chr_genes) {
        if (gene %in% dimnames(interpolated_probs[[chr]])[[3]]) {
          gene_idx <- match(gene, dimnames(interpolated_probs[[chr]])[[3]])
          strain_matrix[gene, ] <- interpolated_probs[[chr]][strain_idx[1], , gene_idx]
        }
      }
    }
  }
  
  return(strain_matrix)
}

# Extract probability matrix from combined parental matrix
extract_combined_probs <- function(combined_matrix, gene_ids) {
  # Create matrix for the combined reference
  combined_strain_matrix <- matrix(0, nrow = length(gene_ids), ncol = 8)
  rownames(combined_strain_matrix) <- gene_ids
  
  # Extract probabilities from the combined matrix
  for (chr in names(combined_matrix)) {
    # Get genes for this chromosome
    chr_genes <- intersect(gene_ids, dimnames(combined_matrix[[chr]])[[3]])
    
    # Extract probabilities for these genes
    for (gene in chr_genes) {
      if (gene %in% dimnames(combined_matrix[[chr]])[[3]]) {
        gene_idx <- match(gene, dimnames(combined_matrix[[chr]])[[3]])
        # Combined matrix has only one row ("Combined")
        combined_strain_matrix[gene, ] <- combined_matrix[[chr]][1, , gene_idx]
      }
    }
  }
  
  return(combined_strain_matrix)
}

# Modified CCRIX sample analysis using only combined parental matrix
analyze_sample_cosine <- function(sample_id, ccrix_base_path, interpolated_probs, 
                                combined_matrix, parent1, parent2) {
  # Construct file path
  sample_file <- file.path(ccrix_base_path, sample_id, 
                          paste0(sample_id, "_gbrs.quantified.diploid.genes.expected_read_counts"))
  
  # Check if file exists
  if (!file.exists(sample_file)) {
    cat("File not found for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Load CCRIX alignment counts
  ccrix_counts <- tryCatch({
    read.table(sample_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  }, error = function(e) {
    cat("Error reading file for sample", sample_id, ":", e$message, "\n")
    return(NULL)
  })
  
  if (is.null(ccrix_counts) || nrow(ccrix_counts) == 0) {
    cat("Empty file for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Check required columns
  required_cols <- c("locus", "total", LETTERS[1:8])
  if (!all(required_cols %in% colnames(ccrix_counts))) {
    cat("Missing required columns for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Convert CCRIX data to proportions matrix
  ccrix_matrix <- ccrix_to_proportions(ccrix_counts)
  
  # Get gene IDs to analyze (those in the CCRIX file with non-zero totals)
  valid_genes <- ccrix_counts$locus[!is.na(ccrix_counts$total) & ccrix_counts$total > 0]
  
  # Extract combined matrix for these genes
  combined_parent_matrix <- extract_combined_probs(combined_matrix, valid_genes)
  
  # Ensure all matrices have the same rows (genes)
  common_genes <- intersect(rownames(ccrix_matrix), rownames(combined_parent_matrix))
  
  if (length(common_genes) == 0) {
    cat("No common genes found for sample", sample_id, "\n")
    return(NULL)
  }
  
  # Add diagnostic info
  cat("Sample", sample_id, "has", length(common_genes), "common genes\n")
  
  # Subset matrices to common genes
  ccrix_matrix <- ccrix_matrix[common_genes, ]
  combined_parent_matrix <- combined_parent_matrix[common_genes, ]
  
  # Remove rows that are all zeros in the ccrix matrix or combined matrix
  non_zero_rows <- rowSums(ccrix_matrix) > 0 & rowSums(combined_parent_matrix) > 0
  
  cat("Sample", sample_id, "has", sum(non_zero_rows), "non-zero rows out of", length(non_zero_rows), "\n")
  
  if (sum(non_zero_rows) == 0) {
    cat("ERROR: No valid non-zero rows found for comparison\n")
    return(NULL)
  }
  
  ccrix_matrix <- ccrix_matrix[non_zero_rows, ]
  combined_parent_matrix <- combined_parent_matrix[non_zero_rows, ]
  
  # Calculate similarity to combined matrix
  sim_to_combined <- cos_sim(ccrix_matrix, combined_parent_matrix)
  
  # For completeness, we'll still extract parent matrices and calculate similarities
  # (even though we don't need to display them)
  valid_genes_filtered <- rownames(ccrix_matrix)
  parent1_matrix <- extract_strain_probs(interpolated_probs, parent1, valid_genes_filtered)
  parent2_matrix <- extract_strain_probs(interpolated_probs, parent2, valid_genes_filtered)
  
  # Calculate similarity to individual parents
  sim_to_parent1 <- cos_sim(ccrix_matrix, parent1_matrix)
  sim_to_parent2 <- cos_sim(ccrix_matrix, parent2_matrix)
  
  # Calculate similarity between parents (as reference)
  parent1_parent2_sim <- cos_sim(parent1_matrix, parent2_matrix)
  
  # Calculate similarity of combined matrix to each parent
  combined_to_parent1 <- cos_sim(combined_parent_matrix, parent1_matrix)
  combined_to_parent2 <- cos_sim(combined_parent_matrix, parent2_matrix)
  
  # Calculate per-gene similarities for detailed analysis
  detailed_sim_parent1 <- cos_sim_per_gene(ccrix_matrix, parent1_matrix)
  detailed_sim_parent2 <- cos_sim_per_gene(ccrix_matrix, parent2_matrix)
  detailed_sim_combined <- cos_sim_per_gene(ccrix_matrix, combined_parent_matrix)
  
  # Calculate gene-level similarities
  gene_similarities <- data.frame(
    gene = rownames(ccrix_matrix),
    sim_parent1 = detailed_sim_parent1$per_gene_sim,
    sim_parent2 = detailed_sim_parent2$per_gene_sim,
    sim_combined = detailed_sim_combined$per_gene_sim,
    stringsAsFactors = FALSE
  )
  
  # Map genes to chromosomes
  chromosomes <- rep("unknown", length(gene_similarities$gene))
  names(chromosomes) <- gene_similarities$gene
  
  # For each chromosome in the interpolated_probs list
  for (chr in names(interpolated_probs)) {
    # Get genes in this chromosome from third dimension of array
    chr_genes <- dimnames(interpolated_probs[[chr]])[[3]]
    
    # Find matching genes between our gene set and this chromosome
    matching_genes <- intersect(gene_similarities$gene, chr_genes)
    
    # Set chromosome for these genes
    if (length(matching_genes) > 0) {
      chromosomes[matching_genes] <- paste0("chr", chr)
    }
  }
  
  # Add chromosome information to the gene_similarities dataframe
  gene_similarities$chromosome <- chromosomes[gene_similarities$gene]
  
  # Calculate per-chromosome similarity
  chr_similarities <- gene_similarities %>%
    dplyr::group_by(chromosome) %>%
    dplyr::summarize(
      sim_combined = mean(sim_combined, na.rm = TRUE),
      sim_parent1 = mean(sim_parent1, na.rm = TRUE),
      sim_parent2 = mean(sim_parent2, na.rm = TRUE),
      gene_count = dplyr::n()
    )
  
  # Create final result object
  result <- list(
    sample_id = sample_id,
    n_genes = nrow(ccrix_matrix),
    sim_to_parent1 = sim_to_parent1,
    sim_to_parent2 = sim_to_parent2,
    sim_to_combined = sim_to_combined,
    parent1_parent2_sim = parent1_parent2_sim,
    combined_to_parent1 = combined_to_parent1,
    combined_to_parent2 = combined_to_parent2,
    
    # Detailed results for further analysis
    gene_similarities = gene_similarities,
    chr_similarities = chr_similarities
  )
  
  return(result)
}

# Modified summarize function with simplified reporting
summarize_cosine_results <- function(cosine_results, parent_strains) {
  summary <- list()
  
  for (cross_type in names(cosine_results)) {
    # Skip if no samples for this cross
    if (length(cosine_results[[cross_type]]) == 0) {
      next
    }
    
    cat("\n=======================================\n")
    cat("Cross:", cross_type, "\n")
    
    # Get parents for this cross
    parents <- parent_strains[[cross_type]]
    parent1 <- parents[1]
    parent2 <- parents[2]
    
    cat("Parental strains:", paste(parents, collapse = " x "), "\n")
    
    # Collect statistics
    sample_count <- length(cosine_results[[cross_type]])
    cat("Total number of samples:", sample_count, "\n")
    
    # Group samples by sex for separate M/F summaries
    samples_by_sex <- list(
      "Female" = names(cosine_results[[cross_type]])[sapply(cosine_results[[cross_type]], function(x) x$sex == "Female")],
      "Male" = names(cosine_results[[cross_type]])[sapply(cosine_results[[cross_type]], function(x) x$sex == "Male")]
    )
    
    cat("Female samples:", length(samples_by_sex$Female), "\n")
    cat("Male samples:", length(samples_by_sex$Male), "\n\n")
    
    # Combine all samples for cross-level results
    all_samples_df <- do.call(rbind, lapply(names(cosine_results[[cross_type]]), function(sample) {
      result <- cosine_results[[cross_type]][[sample]]
      df <- data.frame(
        Sample = sample,
        Sex = result$sex,
        RIX = result$rix,
        Direction = result$cross_direction,
        Matrix = result$matrix_key,
        Genes = result$n_genes,
        Sim_Combined = round(result$sim_to_combined, 3),
        Sim_Parent1 = round(result$sim_to_parent1, 3),
        Sim_Parent2 = round(result$sim_to_parent2, 3),
        stringsAsFactors = FALSE
      )
      return(df)
    }))
    
    # Cross-level summary
    cat("=== OVERALL CROSS SUMMARY ===\n")
    
    # Aggregate results for all samples
    sim_to_parent1 <- sapply(cosine_results[[cross_type]], function(x) x$sim_to_parent1)
    sim_to_parent2 <- sapply(cosine_results[[cross_type]], function(x) x$sim_to_parent2)
    sim_to_combined <- sapply(cosine_results[[cross_type]], function(x) x$sim_to_combined)
    
    # Get parent-parent and combined-parent similarities
    parent1_parent2_sim <- mean(sapply(cosine_results[[cross_type]], function(x) x$parent1_parent2_sim))
    combined_to_parent1 <- mean(sapply(cosine_results[[cross_type]], function(x) x$combined_to_parent1))
    combined_to_parent2 <- mean(sapply(cosine_results[[cross_type]], function(x) x$combined_to_parent2))
    
    genes_analyzed <- sapply(cosine_results[[cross_type]], function(x) x$n_genes)
    
    cat("Total genes analyzed:", sum(genes_analyzed), "\n")
    cat("Average genes per sample:", round(mean(genes_analyzed)), "\n\n")
    
    cat("COSINE SIMILARITY METRICS:\n")
    cat("Average similarity to combined reference:", round(mean(sim_to_combined), 3), "\n")
    cat("Average similarity to", parent1, ":", round(mean(sim_to_parent1), 3), "\n")
    cat("Average similarity to", parent2, ":", round(mean(sim_to_parent2), 3), "\n")
    
    cat("Similarity between", parent1, "and", parent2, ":", round(parent1_parent2_sim, 3), "\n")
    cat("Similarity of combined matrix to", parent1, ":", round(combined_to_parent1, 3), "\n")
    cat("Similarity of combined matrix to", parent2, ":", round(combined_to_parent2, 3), "\n\n")
    
    # Collect chromosome-level summaries
    all_chr_df <- do.call(rbind, lapply(names(cosine_results[[cross_type]]), function(sample) {
      result <- cosine_results[[cross_type]][[sample]]
      if (!is.null(result$chr_similarities)) {
        df <- result$chr_similarities
        df$Sample <- sample
        return(df)
      } else {
        return(NULL)
      }
    }))
    
    # Aggregate chromosome-level results
    if (!is.null(all_chr_df) && nrow(all_chr_df) > 0) {
      chr_summary <- all_chr_df %>%
        dplyr::group_by(chromosome) %>%
        dplyr::summarize(
          avg_sim_combined = mean(sim_combined, na.rm = TRUE),
          sd_sim_combined = sd(sim_combined, na.rm = TRUE),
          avg_genes = mean(gene_count, na.rm = TRUE)
        )
      
      cat("\n=== CHROMOSOME-LEVEL RESULTS ===\n")
      print(chr_summary)
    }
    
    # Store full summary for this cross
    summary[[cross_type]] <- list(
      cross = cross_type,
      parents = parents,
      sample_count = sample_count,
      all_samples = all_samples_df,
      chromosome_summary = chr_summary,
      
      # Cross-level metrics
      avg_sim_combined = mean(sim_to_combined),
      avg_sim_parent1 = mean(sim_to_parent1),
      avg_sim_parent2 = mean(sim_to_parent2),
      parent1_parent2_sim = parent1_parent2_sim,
      combined_to_parent1 = combined_to_parent1,
      combined_to_parent2 = combined_to_parent2,
      total_genes = sum(genes_analyzed),
      avg_genes_per_sample = mean(genes_analyzed)
    )
  }
  
  return(summary)
}

# Modified visualization function to focus only on combined reference and add chromosome similarity
create_visualizations <- function(summary, output_base_dir = "Results/CCRIX") {
  viz_dirs <- list()
  
  for (cross_type in names(summary)) {
    cross_data <- summary[[cross_type]]
    parents <- cross_data$parents
    parent1 <- parents[1]
    parent2 <- parents[2]
    
    # Extract the CC number from the parent string for cleaner labels
    parent1_label <- sub("/.*", "", parent1)
    parent2_label <- sub("/.*", "", parent2)
    
    # Create cross-specific directory structure
    cross_dir <- file.path(output_base_dir, cross_type, "Cosine")
    if (!dir.exists(cross_dir)) {
      dir.create(cross_dir, recursive = TRUE)
    }
    
    # Store the visualization directory for this cross
    viz_dirs[[cross_type]] <- cross_dir
    
    # Create visualizations for all samples
    if (nrow(cross_data$all_samples) > 0) {
      plot_data <- cross_data$all_samples
      
      # 1. Sample-by-sample plot showing ONLY combined reference
      sample_data <- plot_data %>%
        dplyr::select(Sample, Sim_Combined) %>%
        dplyr::mutate(
          Sample = factor(Sample, levels = Sample[order(Sim_Combined, decreasing = TRUE)])
        )
      
      sample_plot_sim <- ggplot(sample_data, aes(x = Sample, y = Sim_Combined)) +
        geom_bar(stat = "identity", fill = "green", alpha = 0.7) +
        theme_minimal() +
        labs(
          title = paste("Cosine Similarity to Combined Reference -", cross_type),
          y = "Cosine Similarity",
          x = "Sample ID"
        ) +
        geom_hline(yintercept = cross_data$avg_sim_combined, 
                   color = "red", linetype = "dashed") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
      
      # Save the sample similarity plot
      sample_sim_file <- file.path(cross_dir, paste0(cross_type, "_sample_similarity.pdf"))
      ggsave(sample_sim_file, sample_plot_sim, width = 12, height = 6)
      
      # 2. Chromosome-level similarity plot
      if (!is.null(cross_data$chromosome_summary) && nrow(cross_data$chromosome_summary) > 0) {
        chr_summary <- cross_data$chromosome_summary
        
        # Sort chromosomes to ensure proper ordering (chr1, chr2, ..., chr19, chrX)
        chr_summary$chromosome <- factor(chr_summary$chromosome, 
                                        levels = c(paste0("chr", 1:19), "chrX"))
        
        chr_summary <- chr_summary %>% dplyr::arrange(chromosome)
        
        # Create chromosome plot
        chr_plot <- ggplot(chr_summary, aes(x = chromosome, y = avg_sim_combined)) +
          geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
          geom_errorbar(aes(ymin = avg_sim_combined - sd_sim_combined, 
                           ymax = avg_sim_combined + sd_sim_combined), width = 0.2) +
          theme_minimal() +
          labs(
            title = paste("Cosine Similarity by Chromosome -", cross_type),
            y = "Similarity to Combined Reference",
            x = "Chromosome"
          ) +
          geom_hline(yintercept = cross_data$avg_sim_combined, 
                    color = "red", linetype = "dashed") +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))
        
        # Save the chromosome plot
        chr_plot_file <- file.path(cross_dir, paste0(cross_type, "_chromosome_similarities.pdf"))
        ggsave(chr_plot_file, chr_plot, width = 10, height = 6)
      }
    }
    
    cat("Visualizations for cross", cross_type, "created in directory:", cross_dir, "\n")
  }
  
  return(viz_dirs)
}

# Add this new function to create the cross-to-RIX mapping
create_cross_to_rix_mapping <- function() {
  # Define the mapping between cross types and RIX
  cross_to_rix <- list()
  
  # RIX1 includes CC001xCC019 and CC019xCC001
  cross_to_rix[["CC001xCC019"]] <- list(rix = "RIX1", direction = "Forward")
  cross_to_rix[["CC019xCC001"]] <- list(rix = "RIX1", direction = "Reverse")
  
  # RIX2 includes CC019xCC040 and CC040xCC019
  cross_to_rix[["CC019xCC040"]] <- list(rix = "RIX2", direction = "Forward")
  cross_to_rix[["CC040xCC019"]] <- list(rix = "RIX2", direction = "Reverse")
  
  # RIX3 includes CC001xCC040 and CC040xCC001
  cross_to_rix[["CC001xCC040"]] <- list(rix = "RIX3", direction = "Forward")
  cross_to_rix[["CC040xCC001"]] <- list(rix = "RIX3", direction = "Reverse")
  
  return(cross_to_rix)
}

# ============================================================================
# FUNCTION TO LOAD COMBINED MATRICES
# ============================================================================

load_combined_matrices <- function(combined_matrices_dir) {
  if (!dir.exists(combined_matrices_dir)) {
    stop("Combined matrices directory not found: ", combined_matrices_dir)
  }
  
  # List all RDS files in the directory
  matrix_files <- list.files(combined_matrices_dir, pattern = ".*_combined_matrix\\.rds$", full.names = TRUE)
  
  if (length(matrix_files) == 0) {
    stop("No combined matrix files found in directory: ", combined_matrices_dir)
  }
  
  cat("Found", length(matrix_files), "combined matrix files\n")
  
  # Load all matrices into a named list
  combined_matrices <- list()
  for (file_path in matrix_files) {
    # Extract the key (RIX_Direction_Sex) from the filename
    filename <- basename(file_path)
    key <- sub("_combined_matrix\\.rds$", "", filename)
    
    cat("Loading matrix:", key, "\n")
    
    # Load the matrix
    matrix_data <- readRDS(file_path)
    combined_matrices[[key]] <- matrix_data
  }
  
  return(combined_matrices)
}

# ============================================================================
# ANALYZE SAMPLES FUNCTION
# ============================================================================

# Function to analyze all samples using the combined matrices

analyze_samples_cosine <- function(
  sample_list,
  ccrix_base_path,
  metadata_subset,
  interpolated_probs,
  combined_matrices,
  parent_strains,
  cross_to_rix,
  target_cross = NULL
) {
  results <- list()
  
  # Initialize results list for each cross type
  for (cross_type in names(parent_strains)) {
    results[[cross_type]] <- list()
  }
  
  # Process each sample
  for (sample_id in sample_list) {
    # Find this sample in the metadata
    sample_row <- which(as.character(metadata_subset$Animal.ID) == sample_id)
    
    if (length(sample_row) == 0) {
      cat("Skipping sample", sample_id, "- not found in metadata\n")
      next
    }
    
    # Get cross type and sex from metadata
    cross_type <- metadata_subset$Cross[sample_row[1]]
    sex <- metadata_subset$Sex[sample_row[1]]
    
    # Skip if targeting a specific cross and this isn't it
    if (!is.null(target_cross) && target_cross != "" && cross_type != target_cross) {
      next
    }
    
    # Skip if cross type not in parent_strains
    if (!(cross_type %in% names(parent_strains))) {
      cat("Skipping sample", sample_id, "- cross type", cross_type, "not defined in parent_strains\n")
      next
    }
    
    # Get parent strains for this cross
    parents <- parent_strains[[cross_type]]
    parent1 <- parents[1]
    parent2 <- parents[2]
    
    # Get RIX and direction information
    rix_info <- cross_to_rix[[cross_type]]
    rix <- rix_info$rix
    direction <- rix_info$direction
    
    # Determine which combined matrix to use
    # Convert sex to proper case format (first letter uppercase, rest lowercase)
    formatted_sex <- paste0(toupper(substr(sex, 1, 1)), tolower(substr(sex, 2, nchar(sex))))
    matrix_key <- paste(rix, direction, formatted_sex, sep = "_")
    
    # Check if we have a combined matrix for this combination
    if (!(matrix_key %in% names(combined_matrices))) {
      cat("WARNING: No combined matrix found for", matrix_key, "\n")
      cat("Available matrices:", paste(names(combined_matrices), collapse = ", "), "\n")
      cat("Trying case-insensitive match...\n")
      
      # Try case-insensitive match
      matrix_names_lower <- tolower(names(combined_matrices))
      matrix_key_lower <- tolower(matrix_key)
      match_idx <- which(matrix_names_lower == matrix_key_lower)
      
      if (length(match_idx) > 0) {
        matrix_key <- names(combined_matrices)[match_idx[1]]
        cat("Found case-insensitive match:", matrix_key, "\n")
      } else {
        cat("Trying to find a matrix for just", rix, "and", direction, "...\n")
        
        # Try to find any matrix for this RIX and direction
        possible_keys <- grep(paste0("^", rix, "_", direction), names(combined_matrices), value = TRUE)
        
        if (length(possible_keys) > 0) {
          matrix_key <- possible_keys[1]
          cat("Using matrix:", matrix_key, "\n")
        } else {
          cat("ERROR: No suitable matrix found for", sample_id, "\n")
          next
        }
      }
    }
    
    cat("Processing sample", sample_id, "(Cross:", cross_type, ", Sex:", sex, ", RIX:", rix, ", Direction:", direction, ")\n")
    
    # Get the combined matrix for this sample
    combined_matrix <- combined_matrices[[matrix_key]]
    
    # Analyze this sample against the combined matrix
    sample_result <- analyze_sample_cosine(
      sample_id = sample_id,
      ccrix_base_path = ccrix_base_path,
      interpolated_probs = interpolated_probs,
      combined_matrix = combined_matrix,
      parent1 = parent1,
      parent2 = parent2
    )
    
    if (!is.null(sample_result)) {
      # Add metadata to the result
      sample_result$sex <- sex
      sample_result$rix <- rix
      sample_result$cross_direction <- direction
      sample_result$matrix_key <- matrix_key
      
      # Store the result
      results[[cross_type]][[sample_id]] <- sample_result
      
      cat("Analyzed", sample_result$n_genes, "genes for sample", sample_id, "\n")
      cat("Cosine similarity to combined reference:", round(sample_result$sim_to_combined, 3), "\n")
    }
  }
  
  return(results)
}

# ============================================================================
# ANALYZE CHROMOSOME SIMILARITIES FUNCTION
# ============================================================================

analyze_chromosome_similarities <- function(results, summary, output_base_dir) {
  for (cross_type in names(results)) {
    cross_samples <- results[[cross_type]]
    
    if (length(cross_samples) == 0) {
      next
    }
    
    # Create cross-specific directory structure
    cross_dir <- file.path(output_base_dir, cross_type, "Cosine")
    if (!dir.exists(cross_dir)) {
      dir.create(cross_dir, recursive = TRUE)
    }
    
    # Extract chromosome data from all samples
    all_chr_data <- do.call(rbind, lapply(names(cross_samples), function(sample_id) {
      sample_result <- cross_samples[[sample_id]]
      
      if (!is.null(sample_result$chr_similarities)) {
        chr_data <- sample_result$chr_similarities
        chr_data$sample_id <- sample_id
        return(chr_data)
      } else {
        return(NULL)
      }
    }))
    
    if (!is.null(all_chr_data) && nrow(all_chr_data) > 0) {
      # Summarize by chromosome
      chr_summary <- all_chr_data %>%
        dplyr::group_by(chromosome) %>%
        dplyr::summarize(
          avg_sim_combined = mean(sim_combined, na.rm = TRUE),
          sd_sim_combined = sd(sim_combined, na.rm = TRUE),
          avg_gene_count = mean(gene_count, na.rm = TRUE),
          sample_count = dplyr::n_distinct(sample_id)
        )
      
      # Save chromosome summary
      chr_summary_file <- file.path(cross_dir, paste0(cross_type, "_chromosome_summary.csv"))
      write.csv(chr_summary, chr_summary_file, row.names = FALSE)
      
      cat("Chromosome similarity summary for", cross_type, "saved to", chr_summary_file, "\n")
    } else {
      cat("No chromosome similarity data available for", cross_type, "\n")
    }
  }
}

# ============================================================================
# RUN CCRIX ANALYSIS FUNCTION
# ============================================================================

run_ccrix_analysis <- function(
  metadata_file,                 # Metadata file path or object
  ccrix_base_path,               # Base path to CCRIX data
  interpolated_probs_file,       # RDS file with interpolated probabilities  
  combined_matrices_dir,         # Directory with combined matrices
  output_base_dir = "Results/CCRIX_Analysis",  # Output directory
  target_cross = NULL            # Optional: Specific cross to analyze
) {
  cat("Loading required data files...\n")
  
  # 1. Load metadata if it's a file path, or use as is if it's already loaded
  if (is.character(metadata_file) && file.exists(metadata_file)) {
    metadata <- read.csv(metadata_file, stringsAsFactors = FALSE)
  } else if (is.character(metadata_file)) {
    stop("Metadata file not found: ", metadata_file)
  } else {
    # Assume it's already loaded as a data frame
    metadata <- metadata_file
  }
  
  # Check metadata has required columns
  required_cols <- c("Animal.ID", "Cross", "Sex")
  missing_cols <- setdiff(required_cols, colnames(metadata))
  if (length(missing_cols) > 0) {
    stop("Metadata is missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  # 2. Load interpolated probabilities
  if (is.character(interpolated_probs_file) && file.exists(interpolated_probs_file)) {
    interpolated_probs <- readRDS(interpolated_probs_file)
  } else if (is.character(interpolated_probs_file) && file.exists(paste0(interpolated_probs_file, ".rds"))) {
    interpolated_probs <- readRDS(paste0(interpolated_probs_file, ".rds"))
  } else if (is.character(interpolated_probs_file)) {
    # Check if it's an object in the environment
    if (exists(interpolated_probs_file)) {
      interpolated_probs <- get(interpolated_probs_file)
    } else {
      stop("Interpolated probabilities file not found: ", interpolated_probs_file)
    }
  } else {
    # Assume it's already loaded
    interpolated_probs <- interpolated_probs_file
  }
  
  # 3. Load combined matrices from directory
  combined_matrices <- load_combined_matrices(combined_matrices_dir)
  
  # 4. Verify the CCRIX base path exists
  if (!dir.exists(ccrix_base_path)) {
    stop("CCRIX base path not found: ", ccrix_base_path)
  }
  
  # 5. Create parent_strains and cross_to_rix mappings from metadata
  
  # Get unique crosses
  crosses <- unique(metadata$Cross)
  
  # For each cross, extract parent strains from the cross name
  parent_strains <- lapply(crosses, function(cross) {
    parts <- strsplit(cross, "x")[[1]]
    if (length(parts) != 2) {
      warning("Invalid cross format: ", cross, ". Expected format: CCXXXxCCYYY")
      return(NULL)
    }
    return(parts)
  })
  names(parent_strains) <- crosses
  
  # Remove any NULL entries (invalid crosses)
  parent_strains <- parent_strains[!sapply(parent_strains, is.null)]
  
  # Use predefined cross_to_rix mapping instead of creating dynamically
  cross_to_rix <- create_cross_to_rix_mapping()
  
  # Check if the requested cross is in our mapping
  if (!is.null(target_cross) && !(target_cross %in% names(cross_to_rix))) {
    stop("Target cross not found in cross_to_rix mapping: ", target_cross)
  }
  
  # 6. Generate sample list from metadata
  sample_list <- as.character(metadata$Animal.ID)
  
  # 7. Filter metadata if target_cross is specified
  if (!is.null(target_cross) && target_cross != "") {
    if (!(target_cross %in% crosses)) {
      stop("Target cross not found in metadata: ", target_cross)
    }
    
    metadata_subset <- metadata[metadata$Cross == target_cross, ]
    cat("Analyzing only cross:", target_cross, "\n")
    cat("Found", nrow(metadata_subset), "samples for this cross\n")
  } else {
    metadata_subset <- metadata
    cat("Analyzing all crosses in metadata\n")
    cat("Total samples:", nrow(metadata_subset), "\n")
  }
  
  # Create output directory if it doesn't exist
  if (!dir.exists(output_base_dir)) {
    dir.create(output_base_dir, recursive = TRUE)
    cat("Created output directory:", output_base_dir, "\n")
  }
  
  # Print summary of what we're going to analyze
  cat("\n==================================================\n")
  cat("CCRIX Analysis Configuration:\n")
  cat("==================================================\n")
  cat("CCRIX data path:", ccrix_base_path, "\n")
  cat("Output directory:", output_base_dir, "\n")
  if (!is.null(target_cross) && target_cross != "") {
    cat("Target cross:", target_cross, "\n")
  } else {
    cat("Crosses to analyze:", paste(names(parent_strains), collapse = ", "), "\n")
  }
  cat("Total samples:", length(sample_list), "\n")
  cat("==================================================\n\n")
  
  # 8. Run the analysis
  cat("Starting CCRIX analysis...\n")
  
  # Run the analysis
  results <- analyze_samples_cosine(
    sample_list = sample_list,
    ccrix_base_path = ccrix_base_path,
    metadata_subset = metadata_subset,
    interpolated_probs = interpolated_probs,
    combined_matrices = combined_matrices,
    parent_strains = parent_strains,
    cross_to_rix = cross_to_rix,
    target_cross = target_cross
  )
  
  # No results generated
  if (length(results) == 0) {
    cat("No results generated. Check your parameters and data.\n")
    return(NULL)
  }
  
  # Summarize results
  cat("\nSummarizing results...\n")
  summary <- summarize_cosine_results(results, parent_strains)
  
  # Create visualizations
  cat("\nCreating visualizations...\n")
  viz_dirs <- create_visualizations(summary, output_base_dir)
  
  # Analyze chromosome-specific similarities
  cat("\nAnalyzing chromosome-specific similarities...\n")
  analyze_chromosome_similarities(results, summary, output_base_dir)
  
  # Generate reports
  cat("\nGenerating reports...\n")
  generate_reports(summary, output_base_dir)
  
  # Save overall results
  if (!is.null(target_cross) && target_cross != "") {
    # When targeting a specific cross, only save results for that cross
    saveRDS(summary, file.path(output_base_dir, paste0(target_cross, "_ccrix_analysis_summary.rds")))
    saveRDS(results, file.path(output_base_dir, paste0(target_cross, "_ccrix_analysis_results.rds")))
    
    # Only save raw results for the target cross
    if (target_cross %in% names(results)) {
      cross_dir <- file.path(output_base_dir, target_cross, "Cosine")
      if (!dir.exists(cross_dir)) {
        dir.create(cross_dir, recursive = TRUE)
      }
      raw_results_file <- file.path(cross_dir, paste0(target_cross, "_cosine_raw_results.rds"))
      saveRDS(results[[target_cross]], raw_results_file)
      cat("Saved raw results for cross", target_cross, "to", raw_results_file, "\n")
    }
  } else {
    # When analyzing all crosses, save everything
    saveRDS(summary, file.path(output_base_dir, "ccrix_analysis_summary.rds"))
    saveRDS(results, file.path(output_base_dir, "ccrix_analysis_results.rds"))
    
    # Save cross-specific raw results files
    for (cross_type in names(results)) {
      # Only save non-empty results
      if (length(results[[cross_type]]) > 0) {
        cross_dir <- file.path(output_base_dir, cross_type, "Cosine")
        if (!dir.exists(cross_dir)) {
          dir.create(cross_dir, recursive = TRUE)
        }
        raw_results_file <- file.path(cross_dir, paste0(cross_type, "_cosine_raw_results.rds"))
        saveRDS(results[[cross_type]], raw_results_file)
        cat("Saved raw results for cross", cross_type, "to", raw_results_file, "\n")
      }
    }
  }
  
  # Final message
  cat("\n==================================================\n")
  if (!is.null(target_cross) && target_cross != "") {
    cat("Analysis for cross", target_cross, "completed successfully!\n")
  } else {
    cat("Analysis for all crosses completed successfully!\n")
  }
  cat("Output files are in:", output_base_dir, "\n")
  cat("==================================================\n")
  
  return(list(results = results, summary = summary))
}


# ============================================================================
# PART 3: RUN THE ANALYSIS
# ============================================================================

# First, define the generate_reports function
generate_reports <- function(summary, output_base_dir) {
  # Generate summary reports for each cross
  for (cross_type in names(summary)) {
    # Create cross-specific directory for output
    cross_dir <- file.path(output_base_dir, cross_type, "Cosine")
    if (!dir.exists(cross_dir)) {
      dir.create(cross_dir, recursive = TRUE)
    }
    
    # Generate report for this cross
    cross_data <- summary[[cross_type]]
    report_file <- file.path(cross_dir, paste0(cross_type, "_cosine_analysis_report.txt"))
    
    sink(report_file)
    cat("CCRIX Cosine Similarity Analysis for", cross_type, "\n")
    cat(paste(rep("=", 50), collapse = ""), "\n\n")
    cat("Analysis Date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
    
    # Write cross-specific information
    cat("\nCross Type:", cross_type, "\n")
    cat("Parent Strains:", paste(cross_data$parents, collapse = " x "), "\n")
    cat("Number of Samples:", cross_data$sample_count, "\n")
    cat("Total Genes Analyzed:", cross_data$total_genes, "\n")
    cat("Average Genes per Sample:", round(cross_data$avg_genes_per_sample), "\n\n")
    
    cat("COSINE SIMILARITY METRICS:\n")
    cat("Average similarity to combined reference:", round(cross_data$avg_sim_combined, 3), "\n")
    
    # Print sample-level results
    cat("\nSAMPLE-LEVEL RESULTS:\n")
    print(cross_data$all_samples[, c("Sample", "Sex", "Genes", "Sim_Combined")])
    
    # Print chromosome-level results if available
    if (!is.null(cross_data$chromosome_summary)) {
      cat("\nCHROMOSOME-LEVEL RESULTS:\n")
      print(cross_data$chromosome_summary)
    }
    
    sink()
    
    cat("Report for cross", cross_type, "generated at", report_file, "\n")
    
    # Also save the summary as an RDS file
    summary_file <- file.path(cross_dir, paste0(cross_type, "_cosine_summary.rds"))
    saveRDS(cross_data, summary_file)
  }
  
  # Create overall summary file
  overall_summary_file <- file.path(output_base_dir, "ccrix_cosine_summary.txt")
  sink(overall_summary_file)
  cat("CCRIX Cosine Similarity Summary\n")
  cat("============================================\n\n")
  cat("Analysis Date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n\n")
  
  # Create a comparison table
  comparison_table <- data.frame(
    Cross = character(),
    Parents = character(),
    Samples = integer(),
    Avg_Similarity = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (cross_type in names(summary)) {
    cross_data <- summary[[cross_type]]
    
    comparison_table <- rbind(comparison_table, data.frame(
      Cross = cross_type,
      Parents = paste(cross_data$parents, collapse = " x "),
      Samples = cross_data$sample_count,
      Avg_Similarity = round(cross_data$avg_sim_combined, 3),
      stringsAsFactors = FALSE
    ))
  }
  
  cat("COMPARISON OF CROSSES:\n")
  print(comparison_table)
  sink()
  
  cat("Overall summary generated at", overall_summary_file, "\n")
}

# Run the analysis with the parameters defined at the top of the script
analysis_results <- run_ccrix_analysis(
  metadata_file = metadata_file,
  ccrix_base_path = ccrix_base_path,
  interpolated_probs_file = interpolated_probs_file,
  combined_matrices_dir = combined_matrices_dir,
  output_base_dir = output_base_dir,
  target_cross = target_cross
)

# Print completion message
cat("\nAnalysis completed successfully. Results are stored in:", output_base_dir, "\n")
```

## 8. Determine/visualize discordant genes

```{r}
# ============================================================================
# Discordant Gene Analysis for CCRIX Data
# ============================================================================

library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)

# ============================================================================
# CONFIGURATION - MODIFY THESE PARAMETERS
# ============================================================================

# Base directory where CCRIX analysis results are stored
results_base_dir <- "Results/CCRIX_G0"

genes_with_id_path <- NULL  # Set to NULL if genes_with_id is already in the environment

# Similarity threshold below which a gene is considered discordant
similarity_threshold <- 0.7

# Output directory for discordant gene analysis
output_base_dir <- "Results/CCRIX_G0_Discordant"

# Optional: Specify a single cross to analyze (e.g., "CC001xCC019")
# Set to NULL to analyze all crosses
target_cross <- NULL

# Bin size for chromosome position binning (in Mb)
bin_size_mb <- 5

# ============================================================================
# HELPER FUNCTIONS
# ============================================================================

# Function to load gene coordinates if not already in environment
load_gene_coordinates <- function(genes_with_id_path) {
  if (!is.null(genes_with_id_path) && file.exists(genes_with_id_path)) {
    return(readRDS(genes_with_id_path))
  } else if (exists("genes_with_id")) {
    return(get("genes_with_id"))
  } else {
    stop("Gene coordinates not found. Please provide a valid path or ensure 'genes_with_id' is loaded.")
  }
}

# Function to load CCRIX analysis results
load_ccrix_results <- function(results_base_dir, target_cross = NULL) {
  # If target_cross is specified, load only that cross's results
  if (!is.null(target_cross)) {
    # Look directly in the cross-specific directory
    cross_dir <- file.path(results_base_dir, target_cross, "Cosine")
    results_file <- file.path(cross_dir, paste0(target_cross, "_cosine_raw_results.rds"))
    
    if (file.exists(results_file)) {
      cat("Loading results for cross", target_cross, "from", results_file, "\n")
      results <- list()
      results[[target_cross]] <- readRDS(results_file)
      return(results)
    } else {
      stop("Results file not found for cross: ", target_cross, " at path: ", results_file)
    }
  } else {
    # Load all crosses
    results <- list()
    cross_dirs <- list.dirs(results_base_dir, full.names = FALSE, recursive = FALSE)
    
    for (cross in cross_dirs) {
      if (grepl("^CC\\d+xCC\\d+$", cross)) {  # Only process directories that look like crosses
        cross_file <- file.path(results_base_dir, cross, "Cosine", paste0(cross, "_cosine_raw_results.rds"))
        if (file.exists(cross_file)) {
          cat("Loading results for cross", cross, "from", cross_file, "\n")
          results[[cross]] <- readRDS(cross_file)
        }
      }
    }
    
    if (length(results) > 0) {
      return(results)
    } else {
      stop("No CCRIX analysis results found in:", results_base_dir)
    }
  }
}

# Function to identify discordant genes for a sample
identify_sample_discordant_genes <- function(sample_result, similarity_threshold) {
  if (is.null(sample_result) || is.null(sample_result$gene_similarities)) {
    return(NULL)
  }
  
  # Get gene similarities
  gene_data <- sample_result$gene_similarities
  
  # Identify discordant genes (those with similarity below threshold)
  discordant_genes <- gene_data %>%
    dplyr::filter(sim_combined < similarity_threshold) %>%
    dplyr::mutate(
      sample_id = sample_result$sample_id
    )
  
  return(discordant_genes)
}

# Function to merge discordant genes with genomic coordinates
merge_discordant_with_coordinates <- function(discordant_genes, gene_coords) {
  # Standardize gene identifiers
  gene_coords$gene_id <- gene_coords$ensembl_gene_id
  
  # For each gene in discordant_genes, find its coordinates
  merged_data <- discordant_genes %>%
    dplyr::left_join(
      gene_coords %>% 
        dplyr::select(gene_id, chromosome_name, start_position, end_position),
      by = c("gene" = "gene_id")
    )
  
  # Calculate gene midpoint position in Mb
  merged_data <- merged_data %>%
    dplyr::mutate(
      midpoint_mb = (start_position + end_position) / 2000000,  # Convert to Mb
      chromosome_name = as.character(chromosome_name)
    )
  
  return(merged_data)
}

# Modified function to calculate concordance percentage by bin
bin_chromosome_positions <- function(merged_data, all_genes_by_chr, bin_size_mb = 5) {
  # First, identify unique discordant genes and their positions
  unique_discordant_genes <- merged_data %>%
    dplyr::select(gene, chromosome_name, midpoint_mb) %>%
    dplyr::distinct()
  
  # Create bins for all chromosomes present in the data
  chromosomes <- unique(unique_discordant_genes$chromosome_name)
  
  # Create a list to store results for each chromosome
  binned_results <- list()
  
  for (chr in chromosomes) {
    # Get all genes for this chromosome from the reference
    chr_all_genes <- all_genes_by_chr[[chr]]
    if (is.null(chr_all_genes) || length(chr_all_genes) == 0) {
      next
    }
    
    # Get discordant genes for this chromosome
    chr_discordant_genes <- unique_discordant_genes %>%
      dplyr::filter(chromosome_name == chr)
    
    # Calculate min and max positions based on all genes in the chromosome
    min_pos <- floor(min(chr_all_genes$midpoint_mb) / bin_size_mb) * bin_size_mb
    max_pos <- ceiling(max(chr_all_genes$midpoint_mb) / bin_size_mb) * bin_size_mb
    
    # Create bins from min to max position
    bins <- seq(min_pos, max_pos, by = bin_size_mb)
    
    # Initialize bin data
    bin_data <- data.frame(
      chromosome_name = chr,
      position_bin = bins,
      position_bin_label = paste0(bins, "-", bins + bin_size_mb),
      total_gene_count = 0,
      discordant_count = 0,
      pct_concordance = 0
    )
    
    # Count total genes in each bin
    for (i in 1:length(bins)) {
      bin_start <- bins[i]
      bin_end <- bin_start + bin_size_mb
      
      # Count all genes in this bin
      bin_data$total_gene_count[i] <- sum(
        chr_all_genes$midpoint_mb >= bin_start & 
        chr_all_genes$midpoint_mb < bin_end
      )
      
      # Count discordant genes in this bin
      bin_data$discordant_count[i] <- sum(
        chr_discordant_genes$midpoint_mb >= bin_start & 
        chr_discordant_genes$midpoint_mb < bin_end
      )
    }
    
    # Calculate concordance percentage (avoiding division by zero)
    bin_data <- bin_data %>%
      dplyr::mutate(
        pct_concordance = ifelse(
          total_gene_count > 0,
          100 * (1 - (discordant_count / total_gene_count)),
          NA
        )
      )
    
    # Store results for this chromosome
    binned_results[[chr]] <- bin_data
  }
  
  # Combine all chromosome results
  binned_data <- do.call(rbind, binned_results)
  
  return(binned_data)
}

# Function to create chromosome position plots with concordance percentage
create_chromosome_concordance_plot <- function(binned_data, chromosome, title = NULL) {
  # Filter data for the specified chromosome
  chr_data <- binned_data %>%
    dplyr::filter(chromosome_name == chromosome) %>%
    dplyr::filter(!is.na(pct_concordance))  # Remove bins with no genes
  
  # If no data for this chromosome, return NULL
  if (nrow(chr_data) == 0) {
    return(NULL)
  }
  
  # Create plot
  plot_title <- ifelse(
    is.null(title),
    paste0("Gene Concordance Distribution - Chromosome ", chromosome),
    title
  )
  
  p <- ggplot(chr_data, aes(x = position_bin, y = pct_concordance)) +
    geom_line(color = "blue", size = 1) +
    geom_point(size = 2, color = "blue") +
    ylim(0, 100) +
    labs(
      title = plot_title,
      x = "Chromosome Position (Mb)",
      y = "% Concordance"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  
  return(p)
}

# Function to analyze discordant genes across all chromosomes
analyze_all_chromosomes <- function(cross_data, gene_coords, similarity_threshold, bin_size_mb,
                                   output_dir) {
  # Create list to store all discordant genes
  all_discordant <- list()
  
  # Process each sample
  for (sample_id in names(cross_data)) {
    cat("Processing sample:", sample_id, "\n")
    sample_result <- cross_data[[sample_id]]
    
    # Identify discordant genes
    discordant_genes <- identify_sample_discordant_genes(sample_result, similarity_threshold)
    
    if (!is.null(discordant_genes) && nrow(discordant_genes) > 0) {
      all_discordant[[sample_id]] <- discordant_genes
    }
  }
  
  # Combine all discordant genes
  if (length(all_discordant) == 0) {
    cat("No discordant genes found for any sample.\n")
    return(NULL)
  }
  
  combined_discordant <- do.call(rbind, all_discordant)
  
  # Merge with genomic coordinates
  merged_data <- merge_discordant_with_coordinates(combined_discordant, gene_coords)
  
  # Save the complete merged data
  full_data_file <- file.path(output_dir, "all_discordant_genes_with_coords.csv")
  write.csv(merged_data, full_data_file, row.names = FALSE)
  cat("Saved all discordant genes with coordinates to:", full_data_file, "\n")
  
  # Filter out entries with missing chromosomes or positions
  valid_data <- merged_data %>% 
    dplyr::filter(!is.na(chromosome_name) & !is.na(midpoint_mb))
  
  if (nrow(valid_data) == 0) {
    cat("No valid chromosome coordinate data found for discordant genes.\n")
    return(NULL)
  }
  
  # Prepare all genes with chromosome positions (for denominator in concordance calculation)
  all_genes_by_chr <- gene_coords %>%
    dplyr::mutate(
      midpoint_mb = (start_position + end_position) / 2000000,  # Convert to Mb
      chromosome_name = as.character(chromosome_name)
    ) %>%
    split(.$chromosome_name)
  
  # Count unique discordant genes by chromosome
  chr_summary <- valid_data %>%
    dplyr::group_by(chromosome_name, gene) %>%
    dplyr::summarize(
      sample_count = n_distinct(sample_id),
      .groups = "drop"
    ) %>%
    dplyr::group_by(chromosome_name) %>%
    dplyr::summarize(
      discordant_count = n(),  # Count unique genes, not sample×gene combinations
      avg_sample_count = mean(sample_count, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(chromosome_name)
  
  # Save chromosome summary
  chr_summary_file <- file.path(output_dir, "chromosome_discordance_summary.csv")
  write.csv(chr_summary, chr_summary_file, row.names = FALSE)
  cat("Saved chromosome discordance summary to:", chr_summary_file, "\n")
  
  # Create binned data for visualization with concordance percentage
  binned_data <- bin_chromosome_positions(valid_data, all_genes_by_chr, bin_size_mb)
  
  # Save binned data
  binned_data_file <- file.path(output_dir, "binned_concordance_data.csv")
  write.csv(binned_data, binned_data_file, row.names = FALSE)
  
  # Create individual chromosome plots
  chromosomes <- unique(valid_data$chromosome_name)
  chromosomes <- sort(chromosomes)  # Sort chromosomes
  
  # Create a directory for chromosome plots
  chr_plots_dir <- file.path(output_dir, "Chromosome_Plots")
  if (!dir.exists(chr_plots_dir)) {
    dir.create(chr_plots_dir, recursive = TRUE)
  }
  
  # Generate individual chromosome plots
  plot_list <- list()
  
  for (chr in chromosomes) {
    cat("Creating concordance plot for chromosome:", chr, "\n")
    
    # Make plot
    chr_plot <- create_chromosome_concordance_plot(binned_data, chr)
    
    if (!is.null(chr_plot)) {
      plot_list[[chr]] <- chr_plot
      
      # Save individual plot as JPEG instead of PDF
      plot_file <- file.path(chr_plots_dir, paste0("concordance_chr", chr, ".jpg"))
      ggsave(plot_file, chr_plot, width = 10, height = 6, device = "jpeg", dpi = 300)
    }
  }
  
  # Create genome-wide summary plot (all chromosomes)
  combined_plot_file <- file.path(output_dir, "genome_wide_concordance.jpg")
  
  if (length(plot_list) > 0) {
    # Arrange multiple plots in a grid
    n_plots <- length(plot_list)
    n_cols <- min(3, n_plots)
    n_rows <- ceiling(n_plots / n_cols)
    
    combined_plot <- gridExtra::grid.arrange(
      grobs = plot_list,
      ncol = n_cols,
      nrow = n_rows,
      top = "Genome-Wide Gene Concordance Distribution"
    )
    
    # Save the combined plot as JPEG
    ggsave(combined_plot_file, combined_plot, width = 18, height = 12, device = "jpeg", dpi = 300)
    cat("Saved genome-wide concordance plot to:", combined_plot_file, "\n")
  } else {
    cat("No chromosome plots were generated.\n")
  }
  
  # Create summary visualizations
  
  # 1. Chromosome bar plot - showing unique gene counts
  chr_summary_plot <- ggplot(chr_summary, aes(x = reorder(chromosome_name, discordant_count), 
                                             y = discordant_count)) +
    geom_bar(stat = "identity", fill = "darkred", alpha = 0.7) +
    geom_text(aes(label = discordant_count), vjust = -0.3) +
    labs(
      title = "Discordant Genes by Chromosome",
      x = "Chromosome",
      y = "Number of Unique Discordant Genes"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  
  # Save chromosome summary plot as JPEG
  chr_summary_plot_file <- file.path(output_dir, "discordant_genes_by_chromosome.jpg")
  ggsave(chr_summary_plot_file, chr_summary_plot, width = 12, height = 8, device = "jpeg", dpi = 300)
  cat("Saved discordant genes by chromosome plot to:", chr_summary_plot_file, "\n")
  
  # 2. Calculate percentage of genes in each sample that are discordant
  sample_gene_counts <- lapply(names(cross_data), function(sample_id) {
    sample_result <- cross_data[[sample_id]]
    if (!is.null(sample_result$gene_similarities)) {
      total_genes <- nrow(sample_result$gene_similarities)
      discordant_count <- sum(sample_result$gene_similarities$sim_combined < similarity_threshold, na.rm = TRUE)
      return(data.frame(
        sample_id = sample_id,
        total_genes = total_genes,
        discordant_genes = discordant_count,
        pct_discordant = (discordant_count / total_genes) * 100
      ))
    } else {
      return(NULL)
    }
  })
  
  # Combine sample gene counts
  sample_gene_counts <- do.call(rbind, sample_gene_counts[!sapply(sample_gene_counts, is.null)])
  
  if (!is.null(sample_gene_counts) && nrow(sample_gene_counts) > 0) {
    # Save sample gene counts
    sample_counts_file <- file.path(output_dir, "sample_discordance_summary.csv")
    write.csv(sample_gene_counts, sample_counts_file, row.names = FALSE)
    
    # Create sample discordance plot
    sample_plot <- ggplot(sample_gene_counts, aes(x = reorder(sample_id, pct_discordant), 
                                                 y = pct_discordant)) +
      geom_bar(stat = "identity", fill = "purple", alpha = 0.7) +
      labs(
        title = "Percentage of Discordant Genes by Sample",
        x = "Sample ID",
        y = "% Discordant Genes"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(hjust = 0.5, face = "bold")
      )
    
    # Save sample plot as JPEG
    sample_plot_file <- file.path(output_dir, "sample_discordance_plot.jpg")
    ggsave(sample_plot_file, sample_plot, width = 12, height = 6, device = "jpeg", dpi = 300)
    cat("Saved sample discordance plot to:", sample_plot_file, "\n")
  }
  
  # Return the processed data for further analysis if needed
  return(list(
    merged_data = merged_data,
    chr_summary = chr_summary,
    binned_data = binned_data,
    sample_summary = sample_gene_counts
  ))
}

# ============================================================================
# MAIN ANALYSIS FUNCTION
# ============================================================================

analyze_discordant_genes <- function(results_base_dir, 
                                    genes_with_id_path,
                                    similarity_threshold, 
                                    output_base_dir,
                                    target_cross = NULL,
                                    bin_size_mb = 5) {
  
  # Load gene coordinates
  cat("Loading gene coordinates...\n")
  gene_coords <- load_gene_coordinates(genes_with_id_path)
  
  # Load CCRIX results
  cat("Loading CCRIX analysis results...\n")
  results <- load_ccrix_results(results_base_dir, target_cross)
  
  # Create output directory if it doesn't exist
  if (!dir.exists(output_base_dir)) {
    dir.create(output_base_dir, recursive = TRUE)
    cat("Created output directory:", output_base_dir, "\n")
  }
  
  # Process each cross
  all_cross_results <- list()
  
  for (cross_type in names(results)) {
    if (length(results[[cross_type]]) == 0) {
      cat("No samples found for cross:", cross_type, "\n")
      next
    }
    
    # Skip if targeting a specific cross and this isn't it
    if (!is.null(target_cross) && target_cross != cross_type) {
      next
    }
    
    cat("\n==================================================\n")
    cat("Analyzing discordant genes for cross:", cross_type, "\n")
    cat("==================================================\n")
    
    # Create cross-specific output directory
    cross_output_dir <- file.path(output_base_dir, cross_type)
    if (!dir.exists(cross_output_dir)) {
      dir.create(cross_output_dir, recursive = TRUE)
    }
    
    # Analyze this cross
    cross_result <- analyze_all_chromosomes(
      cross_data = results[[cross_type]],
      gene_coords = gene_coords,
      similarity_threshold = similarity_threshold,
      bin_size_mb = bin_size_mb,
      output_dir = cross_output_dir
    )
    
    # Store results
    if (!is.null(cross_result)) {
      all_cross_results[[cross_type]] <- cross_result
    }
  }
  
  # If there are multiple crosses, create a cross comparison
  if (length(all_cross_results) > 1) {
    cat("\nCreating cross comparison visualizations...\n")
    
    # Combine chromosome summaries from all crosses
    chr_summaries <- lapply(names(all_cross_results), function(cross) {
      summary <- all_cross_results[[cross]]$chr_summary
      summary$cross <- cross
      return(summary)
    })
    
    combined_chr_summary <- do.call(rbind, chr_summaries)
    
    # Create comparison plot
    comparison_plot <- ggplot(combined_chr_summary, 
                             aes(x = chromosome_name, y = discordant_count, fill = cross)) +
      geom_bar(stat = "identity", position = "dodge") +
      labs(
        title = "Discordant Gene Comparison Across Crosses",
        x = "Chromosome",
        y = "Number of Unique Discordant Genes",
        fill = "Cross"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold")
      )
    
    # Save comparison plot as JPEG
    comparison_plot_file <- file.path(output_base_dir, "cross_comparison_discordant_genes.jpg")
    ggsave(comparison_plot_file, comparison_plot, width = 14, height = 8, device = "jpeg", dpi = 300)
    cat("Saved cross comparison plot to:", comparison_plot_file, "\n")
    
    # Save combined summary
    combined_summary_file <- file.path(output_base_dir, "combined_chromosome_summary.csv")
    write.csv(combined_chr_summary, combined_summary_file, row.names = FALSE)
    
    # Also create a concordance comparison if possible
    if (all(sapply(all_cross_results, function(x) !is.null(x$binned_data)))) {
      # Try to combine binned data for concordance comparison
      binned_data_list <- lapply(names(all_cross_results), function(cross) {
        binned <- all_cross_results[[cross]]$binned_data
        binned$cross <- cross
        return(binned)
      })
      
      combined_binned_data <- do.call(rbind, binned_data_list)
      
      # Save combined binned data
      combined_binned_file <- file.path(output_base_dir, "combined_binned_concordance_data.csv")
      write.csv(combined_binned_data, combined_binned_file, row.names = FALSE)
      
      # For each chromosome, create a comparison plot
      chromosomes <- unique(combined_binned_data$chromosome_name)
      
      # Create directory for comparison plots
      comparison_dir <- file.path(output_base_dir, "Cross_Comparisons")
      if (!dir.exists(comparison_dir)) {
        dir.create(comparison_dir, recursive = TRUE)
      }
      
      for (chr in chromosomes) {
        # Filter data for this chromosome
        chr_data <- combined_binned_data %>%
          dplyr::filter(chromosome_name == chr, !is.na(pct_concordance))
        
        if (nrow(chr_data) > 0) {
          # Create comparison plot
          comp_plot <- ggplot(chr_data, aes(x = position_bin, y = pct_concordance, color = cross, group = cross)) +
            geom_line(size = 1) +
            geom_point(size = 1.5) +
            ylim(0, 100) +
            labs(
              title = paste0("Concordance Comparison - Chromosome ", chr),
              x = "Chromosome Position (Mb)",
              y = "% Concordance",
              color = "Cross"
            ) +
            theme_minimal() +
            theme(
              plot.title = element_text(hjust = 0.5, face = "bold")
            )
          
          # Save comparison plot as JPEG
          comp_plot_file <- file.path(comparison_dir, paste0("concordance_comparison_chr", chr, ".jpg"))
          ggsave(comp_plot_file, comp_plot, width = 12, height = 8, device = "jpeg", dpi = 300)
          cat("Saved concordance comparison for chromosome", chr, "to:", comp_plot_file, "\n")
        }
      }
    }
  }
  
  # Generate a summary report
  cat("\nGenerating summary report...\n")
  
  report_file <- file.path(output_base_dir, "discordant_genes_report.txt")
  sink(report_file)
  
  cat("Discordant Gene Analysis Report\n")
  cat("===============================\n\n")
  cat("Analysis Date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
  cat("Similarity Threshold:", similarity_threshold, "\n")
  cat("Bin Size:", bin_size_mb, "Mb\n\n")
  
  for (cross_type in names(all_cross_results)) {
    cat("\n---------------------------------------\n")
    cat("Cross:", cross_type, "\n")
    cat("---------------------------------------\n")
    
    cross_result <- all_cross_results[[cross_type]]
    
    # Get unique gene count
    unique_genes <- unique(cross_result$merged_data$gene)
    total_discordant <- length(unique_genes)
    
    cat("Total Unique Discordant Genes:", total_discordant, "\n\n")
    
    # Chromosome breakdown
    cat("Chromosome Breakdown:\n")
    print(cross_result$chr_summary[, c("chromosome_name", "discordant_count", "avg_sample_count")])
    
    # Top 5 bins with highest concordance
    if (!is.null(cross_result$binned_data)) {
      top_concordant_bins <- cross_result$binned_data %>%
        dplyr::filter(total_gene_count >= 5) %>%  # Only include bins with at least 5 genes
        dplyr::arrange(desc(pct_concordance)) %>%
        dplyr::slice_head(n = 5)
      
      # Top 5 bins with lowest concordance
      bottom_concordant_bins <- cross_result$binned_data %>%
        dplyr::filter(total_gene_count >= 5) %>%  # Only include bins with at least 5 genes
        dplyr::arrange(pct_concordance) %>%
        dplyr::slice_head(n = 5)
      
      cat("\nTop 5 Regions with Highest Concordance (min 5 genes):\n")
      print(top_concordant_bins[, c("chromosome_name", "position_bin_label", "total_gene_count", "pct_concordance")])
      
      cat("\nTop 5 Regions with Lowest Concordance (min 5 genes):\n")
      print(bottom_concordant_bins[, c("chromosome_name", "position_bin_label", "total_gene_count", "pct_concordance")])
    }
    
    cat("\n")
    
    # Sample statistics - if available
    if (!is.null(cross_result$sample_summary) && nrow(cross_result$sample_summary) > 0) {
      cat("\nSample Summary Statistics:\n")
      cat("Average % of discordant genes per sample:", 
          round(mean(cross_result$sample_summary$pct_discordant), 2), "%\n")
      cat("Range of % discordant genes:", 
          round(min(cross_result$sample_summary$pct_discordant), 2), "% to",
          round(max(cross_result$sample_summary$pct_discordant), 2), "%\n")
    }
  }
  
  sink()
  cat("Summary report generated at:", report_file, "\n")
  
  cat("\n==================================================\n")
  cat("Discordant gene analysis completed successfully!\n")
  cat("Output files are in:", output_base_dir, "\n")
  cat("==================================================\n")
  
  return(all_cross_results)
}

# ============================================================================
# RUN THE ANALYSIS
# ============================================================================

# Execute the analysis with the parameters defined at the top of the script
discordant_results <- analyze_discordant_genes(
  results_base_dir = results_base_dir,
  genes_with_id_path = genes_with_id_path,
  similarity_threshold = similarity_threshold,
  output_base_dir = output_base_dir,
  target_cross = target_cross,
  bin_size_mb = bin_size_mb
)
```

```{r}
# ============================================================================
# Discordant Gene Analysis for CC Data
# ============================================================================

library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)

# ============================================================================
# CONFIGURATION - MODIFY THESE PARAMETERS
# ============================================================================

# Base directory where CCRIX analysis results are stored
results_base_dir <- "Results/CC"

genes_with_id_path <- NULL  # Set to NULL if genes_with_id is already in the environment

# Similarity threshold below which a gene is considered discordant
similarity_threshold <- 0.7

# Output directory for discordant gene analysis
output_base_dir <- "Results/CC_Discordant"

# Optional: Specify a single cross to analyze (e.g., "CC001xCC019")
# Set to NULL to analyze all crosses
target_cross <- NULL

# Bin size for chromosome position binning (in Mb)
bin_size_mb <- 5

# ============================================================================
# HELPER FUNCTIONS
# ============================================================================

# Function to load gene coordinates if not already in environment
load_gene_coordinates <- function(genes_with_id_path) {
  if (!is.null(genes_with_id_path) && file.exists(genes_with_id_path)) {
    return(readRDS(genes_with_id_path))
  } else if (exists("genes_with_id")) {
    return(get("genes_with_id"))
  } else {
    stop("Gene coordinates not found. Please provide a valid path or ensure 'genes_with_id' is loaded.")
  }
}

# Function to load CCRIX analysis results
load_ccrix_results <- function(results_base_dir, target_cross = NULL) {
  # If target_cross is specified, load only that cross's results
  if (!is.null(target_cross)) {
    # Look directly in the cross-specific directory
    cross_dir <- file.path(results_base_dir, target_cross, "Cosine")
    results_file <- file.path(cross_dir, paste0(target_cross, "_cosine_raw_results.rds"))
    
    if (file.exists(results_file)) {
      cat("Loading results for cross", target_cross, "from", results_file, "\n")
      results <- list()
      results[[target_cross]] <- readRDS(results_file)
      return(results)
    } else {
      stop("Results file not found for cross: ", target_cross, " at path: ", results_file)
    }
  } else {
    # Load all crosses
    results <- list()
    cross_dirs <- list.dirs(results_base_dir, full.names = FALSE, recursive = FALSE)
    
    for (cross in cross_dirs) {
      if (grepl("^CC\\d+xCC\\d+$", cross)) {  # Only process directories that look like crosses
        cross_file <- file.path(results_base_dir, cross, "Cosine", paste0(cross, "_cosine_raw_results.rds"))
        if (file.exists(cross_file)) {
          cat("Loading results for cross", cross, "from", cross_file, "\n")
          results[[cross]] <- readRDS(cross_file)
        }
      }
    }
    
    if (length(results) > 0) {
      return(results)
    } else {
      stop("No CCRIX analysis results found in:", results_base_dir)
    }
  }
}

# Modified function to identify discordant genes for a sample - ONLY using sim_strain
identify_sample_discordant_genes <- function(sample_result, similarity_threshold) {
  if (is.null(sample_result) || is.null(sample_result$gene_similarities)) {
    return(NULL)
  }
  
  # Get gene similarities
  gene_data <- sample_result$gene_similarities
  
  # Check if sim_strain exists
  if ("sim_strain" %in% colnames(gene_data)) {
    discordant_genes <- gene_data %>%
      dplyr::filter(sim_strain < similarity_threshold) %>%
      dplyr::mutate(
        sample_id = sample_result$sample_id
      )
  } else {
    cat("'sim_strain' not found in gene similarities data.\n")
    cat("Available columns:", paste(colnames(gene_data), collapse=", "), "\n")
    return(NULL)
  }
  
  return(discordant_genes)
}

# Function to merge discordant genes with genomic coordinates
merge_discordant_with_coordinates <- function(discordant_genes, gene_coords) {
  # Standardize gene identifiers
  gene_coords$gene_id <- gene_coords$ensembl_gene_id
  
  # For each gene in discordant_genes, find its coordinates
  merged_data <- discordant_genes %>%
    dplyr::left_join(
      gene_coords %>% 
        dplyr::select(gene_id, chromosome_name, start_position, end_position),
      by = c("gene" = "gene_id")
    )
  
  # Calculate gene midpoint position in Mb
  merged_data <- merged_data %>%
    dplyr::mutate(
      midpoint_mb = (start_position + end_position) / 2000000,  # Convert to Mb
      chromosome_name = as.character(chromosome_name)
    )
  
  return(merged_data)
}

# Modified function to calculate concordance percentage by bin
bin_chromosome_positions <- function(merged_data, all_genes_by_chr, bin_size_mb = 5) {
  # First, identify unique discordant genes and their positions
  unique_discordant_genes <- merged_data %>%
    dplyr::select(gene, chromosome_name, midpoint_mb) %>%
    dplyr::distinct()
  
  # Create bins for all chromosomes present in the data
  chromosomes <- unique(unique_discordant_genes$chromosome_name)
  
  # Create a list to store results for each chromosome
  binned_results <- list()
  
  for (chr in chromosomes) {
    # Get all genes for this chromosome from the reference
    chr_all_genes <- all_genes_by_chr[[chr]]
    if (is.null(chr_all_genes) || length(chr_all_genes) == 0) {
      next
    }
    
    # Get discordant genes for this chromosome
    chr_discordant_genes <- unique_discordant_genes %>%
      dplyr::filter(chromosome_name == chr)
    
    # Calculate min and max positions based on all genes in the chromosome
    min_pos <- floor(min(chr_all_genes$midpoint_mb) / bin_size_mb) * bin_size_mb
    max_pos <- ceiling(max(chr_all_genes$midpoint_mb) / bin_size_mb) * bin_size_mb
    
    # Create bins from min to max position
    bins <- seq(min_pos, max_pos, by = bin_size_mb)
    
    # Initialize bin data
    bin_data <- data.frame(
      chromosome_name = chr,
      position_bin = bins,
      position_bin_label = paste0(bins, "-", bins + bin_size_mb),
      total_gene_count = 0,
      discordant_count = 0,
      pct_concordance = 0
    )
    
    # Count total genes in each bin
    for (i in 1:length(bins)) {
      bin_start <- bins[i]
      bin_end <- bin_start + bin_size_mb
      
      # Count all genes in this bin
      bin_data$total_gene_count[i] <- sum(
        chr_all_genes$midpoint_mb >= bin_start & 
        chr_all_genes$midpoint_mb < bin_end
      )
      
      # Count discordant genes in this bin
      bin_data$discordant_count[i] <- sum(
        chr_discordant_genes$midpoint_mb >= bin_start & 
        chr_discordant_genes$midpoint_mb < bin_end
      )
    }
    
    # Calculate concordance percentage (avoiding division by zero)
    bin_data <- bin_data %>%
      dplyr::mutate(
        pct_concordance = ifelse(
          total_gene_count > 0,
          100 * (1 - (discordant_count / total_gene_count)),
          NA
        )
      )
    
    # Store results for this chromosome
    binned_results[[chr]] <- bin_data
  }
  
  # Combine all chromosome results
  binned_data <- do.call(rbind, binned_results)
  
  return(binned_data)
}

# Function to create chromosome position plots with concordance percentage
create_chromosome_concordance_plot <- function(binned_data, chromosome, title = NULL) {
  # Filter data for the specified chromosome
  chr_data <- binned_data %>%
    dplyr::filter(chromosome_name == chromosome) %>%
    dplyr::filter(!is.na(pct_concordance))  # Remove bins with no genes
  
  # If no data for this chromosome, return NULL
  if (nrow(chr_data) == 0) {
    return(NULL)
  }
  
  # Create plot
  plot_title <- ifelse(
    is.null(title),
    paste0("Gene Concordance Distribution - Chromosome ", chromosome),
    title
  )
  
  p <- ggplot(chr_data, aes(x = position_bin, y = pct_concordance)) +
    geom_line(color = "blue", size = 1) +
    geom_point(size = 2, color = "blue") +
    ylim(0, 100) +
    labs(
      title = plot_title,
      x = "Chromosome Position (Mb)",
      y = "% Concordance"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  
  return(p)
}

# Function to analyze discordant genes across all chromosomes
analyze_all_chromosomes <- function(cross_data, gene_coords, similarity_threshold, bin_size_mb,
                                   output_dir) {
  # Create list to store all discordant genes
  all_discordant <- list()
  
  # Process each sample
  for (sample_id in names(cross_data)) {
    cat("Processing sample:", sample_id, "\n")
    sample_result <- cross_data[[sample_id]]
    
    # Identify discordant genes
    discordant_genes <- identify_sample_discordant_genes(sample_result, similarity_threshold)
    
    if (!is.null(discordant_genes) && nrow(discordant_genes) > 0) {
      all_discordant[[sample_id]] <- discordant_genes
    }
  }
  
  # Combine all discordant genes
  if (length(all_discordant) == 0) {
    cat("No discordant genes found for any sample.\n")
    return(NULL)
  }
  
  combined_discordant <- do.call(rbind, all_discordant)
  
  # Merge with genomic coordinates
  merged_data <- merge_discordant_with_coordinates(combined_discordant, gene_coords)
  
  # Save the complete merged data
  full_data_file <- file.path(output_dir, "all_discordant_genes_with_coords.csv")
  write.csv(merged_data, full_data_file, row.names = FALSE)
  cat("Saved all discordant genes with coordinates to:", full_data_file, "\n")
  
  # Filter out entries with missing chromosomes or positions
  valid_data <- merged_data %>% 
    dplyr::filter(!is.na(chromosome_name) & !is.na(midpoint_mb))
  
  if (nrow(valid_data) == 0) {
    cat("No valid chromosome coordinate data found for discordant genes.\n")
    return(NULL)
  }
  
  # Prepare all genes with chromosome positions (for denominator in concordance calculation)
  all_genes_by_chr <- gene_coords %>%
    dplyr::mutate(
      midpoint_mb = (start_position + end_position) / 2000000,  # Convert to Mb
      chromosome_name = as.character(chromosome_name)
    ) %>%
    split(.$chromosome_name)
  
  # Count unique discordant genes by chromosome
  chr_summary <- valid_data %>%
    dplyr::group_by(chromosome_name, gene) %>%
    dplyr::summarize(
      sample_count = n_distinct(sample_id),
      .groups = "drop"
    ) %>%
    dplyr::group_by(chromosome_name) %>%
    dplyr::summarize(
      discordant_count = n(),  # Count unique genes, not sample×gene combinations
      avg_sample_count = mean(sample_count, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(chromosome_name)
  
  # Save chromosome summary
  chr_summary_file <- file.path(output_dir, "chromosome_discordance_summary.csv")
  write.csv(chr_summary, chr_summary_file, row.names = FALSE)
  cat("Saved chromosome discordance summary to:", chr_summary_file, "\n")
  
  # Create binned data for visualization with concordance percentage
  binned_data <- bin_chromosome_positions(valid_data, all_genes_by_chr, bin_size_mb)
  
  # Save binned data
  binned_data_file <- file.path(output_dir, "binned_concordance_data.csv")
  write.csv(binned_data, binned_data_file, row.names = FALSE)
  
  # Create individual chromosome plots
  chromosomes <- unique(valid_data$chromosome_name)
  chromosomes <- sort(chromosomes)  # Sort chromosomes
  
  # Create a directory for chromosome plots
  chr_plots_dir <- file.path(output_dir, "Chromosome_Plots")
  if (!dir.exists(chr_plots_dir)) {
    dir.create(chr_plots_dir, recursive = TRUE)
  }
  
  # Generate individual chromosome plots
  plot_list <- list()
  
  for (chr in chromosomes) {
    cat("Creating concordance plot for chromosome:", chr, "\n")
    
    # Make plot
    chr_plot <- create_chromosome_concordance_plot(binned_data, chr)
    
    if (!is.null(chr_plot)) {
      plot_list[[chr]] <- chr_plot
      
      # Save individual plot as JPEG instead of PDF
      plot_file <- file.path(chr_plots_dir, paste0("concordance_chr", chr, ".jpg"))
      ggsave(plot_file, chr_plot, width = 10, height = 6, device = "jpeg", dpi = 300)
    }
  }
  
  # Create genome-wide summary plot (all chromosomes)
  combined_plot_file <- file.path(output_dir, "genome_wide_concordance.jpg")
  
  if (length(plot_list) > 0) {
    # Arrange multiple plots in a grid
    n_plots <- length(plot_list)
    n_cols <- min(3, n_plots)
    n_rows <- ceiling(n_plots / n_cols)
    
    combined_plot <- gridExtra::grid.arrange(
      grobs = plot_list,
      ncol = n_cols,
      nrow = n_rows,
      top = "Genome-Wide Gene Concordance Distribution"
    )
    
    # Save the combined plot as JPEG
    ggsave(combined_plot_file, combined_plot, width = 18, height = 12, device = "jpeg", dpi = 300)
    cat("Saved genome-wide concordance plot to:", combined_plot_file, "\n")
  } else {
    cat("No chromosome plots were generated.\n")
  }
  
  # Create summary visualizations
  
  # 1. Chromosome bar plot - showing unique gene counts
  chr_summary_plot <- ggplot(chr_summary, aes(x = reorder(chromosome_name, discordant_count), 
                                             y = discordant_count)) +
    geom_bar(stat = "identity", fill = "darkred", alpha = 0.7) +
    geom_text(aes(label = discordant_count), vjust = -0.3) +
    labs(
      title = "Discordant Genes by Chromosome",
      x = "Chromosome",
      y = "Number of Unique Discordant Genes"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
  
  # Save chromosome summary plot as JPEG
  chr_summary_plot_file <- file.path(output_dir, "discordant_genes_by_chromosome.jpg")
  ggsave(chr_summary_plot_file, chr_summary_plot, width = 12, height = 8, device = "jpeg", dpi = 300)
  cat("Saved discordant genes by chromosome plot to:", chr_summary_plot_file, "\n")
  
  # 2. Calculate percentage of genes in each sample that are discordant
  sample_gene_counts <- lapply(names(cross_data), function(sample_id) {
    sample_result <- cross_data[[sample_id]]
    if (!is.null(sample_result$gene_similarities)) {
      total_genes <- nrow(sample_result$gene_similarities)
      
      # Check for sim_strain column
      if ("sim_strain" %in% colnames(sample_result$gene_similarities)) {
        discordant_count <- sum(sample_result$gene_similarities$sim_strain < similarity_threshold, na.rm = TRUE)
        
        return(data.frame(
          sample_id = sample_id,
          total_genes = total_genes,
          discordant_genes = discordant_count,
          pct_discordant = (discordant_count / total_genes) * 100
        ))
      } else {
        cat("Warning: 'sim_strain' column not found for sample", sample_id, "\n")
        return(NULL)
      }
    } else {
      return(NULL)
    }
  })
  
  # Combine sample gene counts
  sample_gene_counts <- do.call(rbind, sample_gene_counts[!sapply(sample_gene_counts, is.null)])
  
  if (!is.null(sample_gene_counts) && nrow(sample_gene_counts) > 0) {
    # Save sample gene counts
    sample_counts_file <- file.path(output_dir, "sample_discordance_summary.csv")
    write.csv(sample_gene_counts, sample_counts_file, row.names = FALSE)
    
    # Create sample discordance plot
    sample_plot <- ggplot(sample_gene_counts, aes(x = reorder(sample_id, pct_discordant), 
                                                 y = pct_discordant)) +
      geom_bar(stat = "identity", fill = "purple", alpha = 0.7) +
      labs(
        title = "Percentage of Discordant Genes by Sample",
        x = "Sample ID",
        y = "% Discordant Genes"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(hjust = 0.5, face = "bold")
      )
    
    # Save sample plot as JPEG
    sample_plot_file <- file.path(output_dir, "sample_discordance_plot.jpg")
    ggsave(sample_plot_file, sample_plot, width = 12, height = 6, device = "jpeg", dpi = 300)
    cat("Saved sample discordance plot to:", sample_plot_file, "\n")
  }
  
  # Return the processed data for further analysis if needed
  return(list(
    merged_data = merged_data,
    chr_summary = chr_summary,
    binned_data = binned_data,
    sample_summary = sample_gene_counts
  ))
}

# ============================================================================
# MAIN ANALYSIS FUNCTION
# ============================================================================

analyze_discordant_genes <- function(results_base_dir, 
                                    genes_with_id_path,
                                    similarity_threshold, 
                                    output_base_dir,
                                    target_cross = NULL,
                                    bin_size_mb = 5) {
  
  # Load gene coordinates
  cat("Loading gene coordinates...\n")
  gene_coords <- load_gene_coordinates(genes_with_id_path)
  
  # Load CCRIX results
  cat("Loading CCRIX analysis results...\n")
  results <- load_ccrix_results(results_base_dir, target_cross)
  
  # Create output directory if it doesn't exist
  if (!dir.exists(output_base_dir)) {
    dir.create(output_base_dir, recursive = TRUE)
    cat("Created output directory:", output_base_dir, "\n")
  }
  
  # Process each cross
  all_cross_results <- list()
  
  for (cross_type in names(results)) {
    if (length(results[[cross_type]]) == 0) {
      cat("No samples found for cross:", cross_type, "\n")
      next
    }
    
    # Skip if targeting a specific cross and this isn't it
    if (!is.null(target_cross) && target_cross != cross_type) {
      next
    }
    
    cat("\n==================================================\n")
    cat("Analyzing discordant genes for cross:", cross_type, "\n")
    cat("==================================================\n")
    
    # Create cross-specific output directory
    cross_output_dir <- file.path(output_base_dir, cross_type)
    if (!dir.exists(cross_output_dir)) {
      dir.create(cross_output_dir, recursive = TRUE)
    }
    
    # Analyze this cross
    cross_result <- analyze_all_chromosomes(
      cross_data = results[[cross_type]],
      gene_coords = gene_coords,
      similarity_threshold = similarity_threshold,
      bin_size_mb = bin_size_mb,
      output_dir = cross_output_dir
    )
    
    # Store results
    if (!is.null(cross_result)) {
      all_cross_results[[cross_type]] <- cross_result
    }
  }
  
  # If there are multiple crosses, create a cross comparison
  if (length(all_cross_results) > 1) {
    cat("\nCreating cross comparison visualizations...\n")
    
    # Combine chromosome summaries from all crosses
    chr_summaries <- lapply(names(all_cross_results), function(cross) {
      summary <- all_cross_results[[cross]]$chr_summary
      summary$cross <- cross
      return(summary)
    })
    
    combined_chr_summary <- do.call(rbind, chr_summaries)
    
    # Create comparison plot
    comparison_plot <- ggplot(combined_chr_summary, 
                             aes(x = chromosome_name, y = discordant_count, fill = cross)) +
      geom_bar(stat = "identity", position = "dodge") +
      labs(
        title = "Discordant Gene Comparison Across Crosses",
        x = "Chromosome",
        y = "Number of Unique Discordant Genes",
        fill = "Cross"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold")
      )
    
    # Save comparison plot as JPEG
    comparison_plot_file <- file.path(output_base_dir, "cross_comparison_discordant_genes.jpg")
    ggsave(comparison_plot_file, comparison_plot, width = 14, height = 8, device = "jpeg", dpi = 300)
    cat("Saved cross comparison plot to:", comparison_plot_file, "\n")
    
    # Save combined summary
    combined_summary_file <- file.path(output_base_dir, "combined_chromosome_summary.csv")
    write.csv(combined_chr_summary, combined_summary_file, row.names = FALSE)
    
    # Also create a concordance comparison if possible
    if (all(sapply(all_cross_results, function(x) !is.null(x$binned_data)))) {
      # Try to combine binned data for concordance comparison
      binned_data_list <- lapply(names(all_cross_results), function(cross) {
        binned <- all_cross_results[[cross]]$binned_data
        binned$cross <- cross
        return(binned)
      })
      
      combined_binned_data <- do.call(rbind, binned_data_list)
      
      # Save combined binned data
      combined_binned_file <- file.path(output_base_dir, "combined_binned_concordance_data.csv")
      write.csv(combined_binned_data, combined_binned_file, row.names = FALSE)
      
      # For each chromosome, create a comparison plot
      chromosomes <- unique(combined_binned_data$chromosome_name)
      
      # Create directory for comparison plots
      comparison_dir <- file.path(output_base_dir, "Cross_Comparisons")
      if (!dir.exists(comparison_dir)) {
        dir.create(comparison_dir, recursive = TRUE)
      }
      
      for (chr in chromosomes) {
        # Filter data for this chromosome
        chr_data <- combined_binned_data %>%
          dplyr::filter(chromosome_name == chr, !is.na(pct_concordance))
        
        if (nrow(chr_data) > 0) {
          # Create comparison plot
          comp_plot <- ggplot(chr_data, aes(x = position_bin, y = pct_concordance, color = cross, group = cross)) +
            geom_line(size = 1) +
            geom_point(size = 1.5) +
            ylim(0, 100) +
            labs(
              title = paste0("Concordance Comparison - Chromosome ", chr),
              x = "Chromosome Position (Mb)",
              y = "% Concordance",
              color = "Cross"
            ) +
            theme_minimal() +
            theme(
              plot.title = element_text(hjust = 0.5, face = "bold")
            )
          
          # Save comparison plot as JPEG
          comp_plot_file <- file.path(comparison_dir, paste0("concordance_comparison_chr", chr, ".jpg"))
          ggsave(comp_plot_file, comp_plot, width = 12, height = 8, device = "jpeg", dpi = 300)
          cat("Saved concordance comparison for chromosome", chr, "to:", comp_plot_file, "\n")
        }
      }
    }
  }
  
  # Generate a summary report
  cat("\nGenerating summary report...\n")
  
  report_file <- file.path(output_base_dir, "discordant_genes_report.txt")
  sink(report_file)
  
  cat("Discordant Gene Analysis Report\n")
  cat("===============================\n\n")
  cat("Analysis Date:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
  cat("Similarity Threshold:", similarity_threshold, "\n")
  cat("Bin Size:", bin_size_mb, "Mb\n\n")
  
  for (cross_type in names(all_cross_results)) {
    cat("\n---------------------------------------\n")
    cat("Cross:", cross_type, "\n")
    cat("---------------------------------------\n")
    
    cross_result <- all_cross_results[[cross_type]]
    
    # Get unique gene count
    unique_genes <- unique(cross_result$merged_data$gene)
    total_discordant <- length(unique_genes)
    
    cat("Total Unique Discordant Genes:", total_discordant, "\n\n")
    
    # Chromosome breakdown
    cat("Chromosome Breakdown:\n")
    print(cross_result$chr_summary[, c("chromosome_name", "discordant_count", "avg_sample_count")])
    
    # Top 5 bins with highest concordance
    if (!is.null(cross_result$binned_data)) {
      top_concordant_bins <- cross_result$binned_data %>%
        dplyr::filter(total_gene_count >= 5) %>%  # Only include bins with at least 5 genes
        dplyr::arrange(desc(pct_concordance)) %>%
        dplyr::slice_head(n = 5)
      
      # Top 5 bins with lowest concordance
      bottom_concordant_bins <- cross_result$binned_data %>%
        dplyr::filter(total_gene_count >= 5) %>%  # Only include bins with at least 5 genes
        dplyr::arrange(pct_concordance) %>%
        dplyr::slice_head(n = 5)
      
      cat("\nTop 5 Regions with Highest Concordance (min 5 genes):\n")
      print(top_concordant_bins[, c("chromosome_name", "position_bin_label", "total_gene_count", "pct_concordance")])
      
      cat("\nTop 5 Regions with Lowest Concordance (min 5 genes):\n")
      print(bottom_concordant_bins[, c("chromosome_name", "position_bin_label", "total_gene_count", "pct_concordance")])
    }
    
    cat("\n")
    
    # Sample statistics - if available
    if (!is.null(cross_result$sample_summary) && nrow(cross_result$sample_summary) > 0) {
      cat("\nSample Summary Statistics:\n")
      cat("Average % of discordant genes per sample:", 
          round(mean(cross_result$sample_summary$pct_discordant), 2), "%\n")
      cat("Range of % discordant genes:", 
          round(min(cross_result$sample_summary$pct_discordant), 2), "% to",
          round(max(cross_result$sample_summary$pct_discordant), 2), "%\n")
    }
  }
  
  sink()
  cat("Summary report generated at:", report_file, "\n")
  
  cat("\n==================================================\n")
  cat("Discordant gene analysis completed successfully!\n")
  cat("Output files are in:", output_base_dir, "\n")
  cat("==================================================\n")
  
  return(all_cross_results)
}

# ============================================================================
# RUN THE ANALYSIS
# ============================================================================

# Execute the analysis with the parameters defined at the top of the script
discordant_results <- analyze_discordant_genes(
  results_base_dir = results_base_dir,
  genes_with_id_path = genes_with_id_path,
  similarity_threshold = similarity_threshold,
  output_base_dir = output_base_dir,
  target_cross = target_cross,
  bin_size_mb = bin_size_mb
)
```

```{r}
# Visualize CCRIX with parental CC strains 

# ============================================================================
# Simplified Discordant Gene Analysis for CC Data - Concordance Comparisons Only
# ============================================================================

library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)
library(cowplot)  # For get_legend function

# ============================================================================
# CONFIGURATION - MODIFY THESE PARAMETERS
# ============================================================================

# Base directory where CC analysis results are stored
cc_results_dir <- "Results/CC"

# Base directory where CCRIX G0 analysis results are stored
ccrix_results_dir <- "Results/CCRIX_G0"

genes_with_id_path <- NULL  # Set to NULL if genes_with_id is already in the environment

# Similarity threshold below which a gene is considered discordant
similarity_threshold <- 0.7

# Output directory for discordant gene analysis
output_base_dir <- "Results/CC_CCRIX_Comparison"

# Bin size for chromosome position binning (in Mb)
bin_size_mb <- 5

# ============================================================================
# HELPER FUNCTIONS
# ============================================================================

# Function to load gene coordinates if not already in environment
load_gene_coordinates <- function(genes_with_id_path) {
  if (!is.null(genes_with_id_path) && file.exists(genes_with_id_path)) {
    return(readRDS(genes_with_id_path))
  } else if (exists("genes_with_id")) {
    return(get("genes_with_id"))
  } else {
    stop("Gene coordinates not found. Please provide a valid path or ensure 'genes_with_id' is loaded.")
  }
}

# Function to load CC analysis results
load_cc_results <- function(results_base_dir) {
  results <- list()
  cross_dirs <- list.dirs(results_base_dir, full.names = FALSE, recursive = FALSE)
  
  for (cross in cross_dirs) {
    if (grepl("^CC\\d+xCC\\d+$", cross)) {  # Only process directories that look like crosses
      cross_file <- file.path(results_base_dir, cross, "Cosine", paste0(cross, "_cosine_raw_results.rds"))
      if (file.exists(cross_file)) {
        cat("Loading results for cross", cross, "from", cross_file, "\n")
        results[[cross]] <- readRDS(cross_file)
      }
    }
  }
  
  if (length(results) > 0) {
    return(results)
  } else {
    stop("No CC analysis results found in:", results_base_dir)
  }
}

# Function to load CCRIX G0 analysis results
load_ccrix_results <- function(results_base_dir) {
  results <- list()
  cross_dirs <- list.dirs(results_base_dir, full.names = FALSE, recursive = FALSE)
  
  for (cross in cross_dirs) {
    if (grepl("^CC\\d+xCC\\d+$", cross)) {  # Only process directories that look like crosses
      cross_file <- file.path(results_base_dir, cross, "Cosine", paste0(cross, "_cosine_raw_results.rds"))
      if (file.exists(cross_file)) {
        cat("Loading results for CCRIX cross", cross, "from", cross_file, "\n")
        results[[cross]] <- readRDS(cross_file)
      }
    }
  }
  
  if (length(results) > 0) {
    return(results)
  } else {
    stop("No CCRIX G0 analysis results found in:", results_base_dir)
  }
}

# Modified function to identify discordant genes for a sample - handles both CC and CCRIX formats
identify_sample_discordant_genes <- function(sample_result, similarity_threshold, data_type = "CC") {
  if (is.null(sample_result) || is.null(sample_result$gene_similarities)) {
    return(NULL)
  }
  
  # Get gene similarities
  gene_data <- sample_result$gene_similarities
  
  # For CC data, use sim_strain column
  if (data_type == "CC" && "sim_strain" %in% colnames(gene_data)) {
    discordant_genes <- gene_data %>%
      dplyr::filter(sim_strain < similarity_threshold) %>%
      dplyr::mutate(
        sample_id = sample_result$sample_id
      )
    return(discordant_genes)
  }
  # For CCRIX data, use sim_combined column
  else if (data_type == "CCRIX" && "sim_combined" %in% colnames(gene_data)) {
    discordant_genes <- gene_data %>%
      dplyr::filter(sim_combined < similarity_threshold) %>%
      dplyr::mutate(
        sample_id = sample_result$sample_id
      )
    return(discordant_genes)
  }
  else {
    available_cols <- paste(colnames(gene_data), collapse=", ")
    cat("Required similarity column not found in gene similarities data.\n")
    cat("Available columns:", available_cols, "\n")
    return(NULL)
  }
}

# Function to merge discordant genes with genomic coordinates
merge_discordant_with_coordinates <- function(discordant_genes, gene_coords) {
  # Standardize gene identifiers
  gene_coords$gene_id <- gene_coords$ensembl_gene_id
  
  # For each gene in discordant_genes, find its coordinates
  merged_data <- discordant_genes %>%
    dplyr::left_join(
      gene_coords %>% 
        dplyr::select(gene_id, chromosome_name, start_position, end_position),
      by = c("gene" = "gene_id")
    )
  
  # Calculate gene midpoint position in Mb
  merged_data <- merged_data %>%
    dplyr::mutate(
      midpoint_mb = (start_position + end_position) / 2000000,  # Convert to Mb
      chromosome_name = as.character(chromosome_name)
    )
  
  return(merged_data)
}

# Function to calculate concordance percentage by bin
bin_chromosome_positions <- function(merged_data, all_genes_by_chr, bin_size_mb = 5) {
  # First, identify unique discordant genes and their positions
  unique_discordant_genes <- merged_data %>%
    dplyr::select(gene, chromosome_name, midpoint_mb) %>%
    dplyr::distinct()
  
  # Create bins for all chromosomes present in the data
  chromosomes <- unique(unique_discordant_genes$chromosome_name)
  
  # Create a list to store results for each chromosome
  binned_results <- list()
  
  for (chr in chromosomes) {
    # Get all genes for this chromosome from the reference
    chr_all_genes <- all_genes_by_chr[[chr]]
    if (is.null(chr_all_genes) || length(chr_all_genes) == 0) {
      next
    }
    
    # Get discordant genes for this chromosome
    chr_discordant_genes <- unique_discordant_genes %>%
      dplyr::filter(chromosome_name == chr)
    
    # Calculate min and max positions based on all genes in the chromosome
    min_pos <- floor(min(chr_all_genes$midpoint_mb) / bin_size_mb) * bin_size_mb
    max_pos <- ceiling(max(chr_all_genes$midpoint_mb) / bin_size_mb) * bin_size_mb
    
    # Create bins from min to max position
    bins <- seq(min_pos, max_pos, by = bin_size_mb)
    
    # Initialize bin data
    bin_data <- data.frame(
      chromosome_name = chr,
      position_bin = bins,
      position_bin_label = paste0(bins, "-", bins + bin_size_mb),
      total_gene_count = 0,
      discordant_count = 0,
      pct_concordance = 0
    )
    
    # Count total genes in each bin
    for (i in 1:length(bins)) {
      bin_start <- bins[i]
      bin_end <- bin_start + bin_size_mb
      
      # Count all genes in this bin
      bin_data$total_gene_count[i] <- sum(
        chr_all_genes$midpoint_mb >= bin_start & 
        chr_all_genes$midpoint_mb < bin_end
      )
      
      # Count discordant genes in this bin
      bin_data$discordant_count[i] <- sum(
        chr_discordant_genes$midpoint_mb >= bin_start & 
        chr_discordant_genes$midpoint_mb < bin_end
      )
    }
    
    # Calculate concordance percentage (avoiding division by zero)
    bin_data <- bin_data %>%
      dplyr::mutate(
        pct_concordance = ifelse(
          total_gene_count > 0,
          100 * (1 - (discordant_count / total_gene_count)),
          NA
        )
      )
    
    # Store results for this chromosome
    binned_results[[chr]] <- bin_data
  }
  
  # Combine all chromosome results
  if (length(binned_results) > 0) {
    binned_data <- do.call(rbind, binned_results)
    return(binned_data)
  } else {
    return(NULL)
  }
}

# Improved function to extract strain IDs from cross name
extract_strains <- function(cross_name) {
  # Match the pattern CCxxxCCyyy
  matches <- regmatches(cross_name, regexec("CC(\\d+)xCC(\\d+)", cross_name))
  if (length(matches) > 0 && length(matches[[1]]) > 2) {
    strain1 <- as.numeric(matches[[1]][2])
    strain2 <- as.numeric(matches[[1]][3])
    return(c(strain1, strain2))
  }
  cat("Warning: Could not extract strains from cross name:", cross_name, "\n")
  return(NULL)
}

# Function to process data and create concordance comparison visualizations
create_concordance_comparisons <- function(cc_results, ccrix_results, gene_coords, 
                                          similarity_threshold, bin_size_mb, output_dir) {
  
  # Define consistent colors for parent crosses and CCRIX
  color_scheme <- c(
    "CC001xCC001" = "red",
    "CC019xCC019" = "blue",
    "CC040xCC040" = "green",
    "CCRIX" = "yellow"
  )
  
  # Process CC results
  cc_processed <- list()
  for (cross in names(cc_results)) {
    cat("\nProcessing CC cross:", cross, "\n")
    
    # Extract all discordant genes for this cross
    all_discordant <- list()
    for (sample_id in names(cc_results[[cross]])) {
      sample_result <- cc_results[[cross]][[sample_id]]
      discordant_genes <- identify_sample_discordant_genes(sample_result, similarity_threshold, data_type = "CC")
      if (!is.null(discordant_genes) && nrow(discordant_genes) > 0) {
        all_discordant[[sample_id]] <- discordant_genes
      }
    }
    
    if (length(all_discordant) == 0) {
      cat("No discordant genes found for cross:", cross, "\n")
      next
    }
    
    # Combine all discordant genes
    combined_discordant <- do.call(rbind, all_discordant)
    
    # Add cross identifier
    combined_discordant$cross <- cross
    combined_discordant$data_type <- "CC"
    
    # Extract strain IDs
    strains <- extract_strains(cross)
    if (!is.null(strains)) {
      combined_discordant$strain1 <- strains[1]
      combined_discordant$strain2 <- strains[2]
    }
    
    # Merge with genomic coordinates
    merged_data <- merge_discordant_with_coordinates(combined_discordant, gene_coords)
    
    # Store for comparison
    cc_processed[[cross]] <- merged_data
  }
  
  # Process CCRIX results
  ccrix_processed <- list()
  for (cross in names(ccrix_results)) {
    cat("\nProcessing CCRIX cross:", cross, "\n")
    
    # Extract strain IDs
    strains <- extract_strains(cross)
    if (is.null(strains)) {
      cat("Could not extract strains from cross name:", cross, "\n")
      next
    }
    
    # Extract all discordant genes for this cross
    all_discordant <- list()
    for (sample_id in names(ccrix_results[[cross]])) {
      sample_result <- ccrix_results[[cross]][[sample_id]]
      discordant_genes <- identify_sample_discordant_genes(sample_result, similarity_threshold, data_type = "CCRIX")
      if (!is.null(discordant_genes) && nrow(discordant_genes) > 0) {
        all_discordant[[sample_id]] <- discordant_genes
      }
    }
    
    if (length(all_discordant) == 0) {
      cat("No discordant genes found for CCRIX cross:", cross, "\n")
      next
    }
    
    # Combine all discordant genes
    combined_discordant <- do.call(rbind, all_discordant)
    
    # Add cross identifier
    combined_discordant$cross <- cross
    combined_discordant$data_type <- "CCRIX"
    combined_discordant$strain1 <- strains[1]
    combined_discordant$strain2 <- strains[2]
    
    # Merge with genomic coordinates
    merged_data <- merge_discordant_with_coordinates(combined_discordant, gene_coords)
    
    # Store for comparison
    ccrix_processed[[cross]] <- merged_data
  }
  
  # Function to find parent crosses for a CCRIX cross
  find_parent_crosses <- function(ccrix_cross, cc_crosses) {
    strains <- extract_strains(ccrix_cross)
    if (is.null(strains)) return(NULL)
    
    strain1 <- strains[1]
    strain2 <- strains[2]
    
    # Look for parent crosses in the format CC001xCC001 and CC019xCC019 (with leading zeros)
    parent1_cross <- sprintf("CC%03dxCC%03d", strain1, strain1)
    parent2_cross <- sprintf("CC%03dxCC%03d", strain2, strain2)
    
    # Check if both parent crosses exist in cc_crosses
    parents <- c()
    if (parent1_cross %in% names(cc_crosses)) {
      parents <- c(parents, parent1_cross)
    } else {
      cat("Warning: Parent cross", parent1_cross, "not found for CCRIX cross", ccrix_cross, "\n")
    }
    
    if (parent2_cross %in% names(cc_crosses)) {
      parents <- c(parents, parent2_cross)
    } else {
      cat("Warning: Parent cross", parent2_cross, "not found for CCRIX cross", ccrix_cross, "\n")
    }
    
    if (length(parents) > 0) {
      return(parents)
    } else {
      return(NULL)
    }
  }
  
  # Create comparisons between CCRIX and parent CC data
  if (length(cc_processed) > 0 && length(ccrix_processed) > 0) {
    cat("\nCreating CCRIX and parent CC strain comparisons...\n")
    
    # Create base output directory for comparisons
    base_comparison_dir <- file.path(output_dir, "Cross_Comparisons")
    if (!dir.exists(base_comparison_dir)) {
      dir.create(base_comparison_dir, recursive = TRUE)
    }
    
    # Prepare all genes with chromosome positions (for denominator in concordance calculation)
    all_genes_by_chr <- gene_coords %>%
      dplyr::mutate(
        midpoint_mb = (start_position + end_position) / 2000000,  # Convert to Mb
        chromosome_name = as.character(chromosome_name)
      ) %>%
      split(.$chromosome_name)
    
    # Process each CCRIX cross
    for (ccrix_cross in names(ccrix_processed)) {
      cat("\nProcessing comparison for CCRIX cross:", ccrix_cross, "\n")
      
      # Create cross-specific output directory
      comparison_dir <- file.path(base_comparison_dir, ccrix_cross)
      if (!dir.exists(comparison_dir)) {
        dir.create(comparison_dir, recursive = TRUE)
      }
      
      # Find parent crosses
      parent_crosses <- find_parent_crosses(ccrix_cross, cc_processed)
      
      if (is.null(parent_crosses) || length(parent_crosses) == 0) {
        cat("No parent crosses found for CCRIX cross:", ccrix_cross, "\n")
        next
      }
      
      cat("Found parent crosses:", paste(parent_crosses, collapse=", "), "\n")
      
      # Extract CCRIX strains for labeling
      ccrix_strains <- extract_strains(ccrix_cross)
      if (is.null(ccrix_strains)) next
      
      strain1 <- paste0("CC", sprintf("%03d", ccrix_strains[1]))
      strain2 <- paste0("CC", sprintf("%03d", ccrix_strains[2]))
      
      # Get the combined columns from all datasets to ensure consistency
      ccrix_data <- ccrix_processed[[ccrix_cross]]
      parent1_data <- cc_processed[[parent_crosses[1]]]
      parent2_data <- cc_processed[[parent_crosses[2]]]
      
      # Get a list of all column names across all three dataframes
      all_cols <- unique(c(
        colnames(ccrix_data), 
        colnames(parent1_data), 
        colnames(parent2_data)
      ))
      
      # Function to ensure a dataframe has all necessary columns
      ensure_columns <- function(df, required_cols) {
        missing_cols <- setdiff(required_cols, colnames(df))
        if (length(missing_cols) > 0) {
          for (col in missing_cols) {
            df[[col]] <- NA
          }
        }
        return(df[, required_cols])
      }
      
      # Ensure all dataframes have the same columns
      ccrix_data <- ensure_columns(ccrix_data, all_cols)
      parent1_data <- ensure_columns(parent1_data, all_cols)
      parent2_data <- ensure_columns(parent2_data, all_cols)
      
      # Now combine data with consistent columns
      combined_data <- rbind(
        ccrix_data %>% mutate(group = "CCRIX", label = paste0(strain1, "x", strain2)),
        parent1_data %>% mutate(group = "Parent1", label = parent_crosses[1]),
        parent2_data %>% mutate(group = "Parent2", label = parent_crosses[2])
      )
      
      # Only keep valid data with chromosome and position information
      valid_data <- combined_data %>% 
        dplyr::filter(!is.na(chromosome_name) & !is.na(midpoint_mb))
      
      if (nrow(valid_data) == 0) {
        cat("No valid chromosome coordinate data found for cross comparison.\n")
        next
      }
      
      # Calculate binned data for each group
      ccrix_data <- valid_data %>% filter(group == "CCRIX")
      parent1_data <- valid_data %>% filter(group == "Parent1")
      parent2_data <- valid_data %>% filter(group == "Parent2")
      
      ccrix_binned <- bin_chromosome_positions(ccrix_data, all_genes_by_chr, bin_size_mb)
      if (!is.null(ccrix_binned)) {
        ccrix_binned$group <- "CCRIX"
        ccrix_binned$label <- paste0(strain1, "x", strain2)
      } else {
        cat("No binned data could be generated for CCRIX cross:", ccrix_cross, "\n")
        next
      }
      
      parent1_binned <- bin_chromosome_positions(parent1_data, all_genes_by_chr, bin_size_mb)
      if (!is.null(parent1_binned)) {
        parent1_binned$group <- "Parent1"
        parent1_binned$label <- parent_crosses[1]
      } else {
        cat("No binned data could be generated for parent cross:", parent_crosses[1], "\n")
        next
      }
      
      parent2_binned <- bin_chromosome_positions(parent2_data, all_genes_by_chr, bin_size_mb)
      if (!is.null(parent2_binned)) {
        parent2_binned$group <- "Parent2"
        parent2_binned$label <- parent_crosses[2]
      } else {
        cat("No binned data could be generated for parent cross:", parent_crosses[2], "\n")
        next
      }
      
      # Combine binned data
      combined_binned <- rbind(ccrix_binned, parent1_binned, parent2_binned)
      
      # Save combined binned data
      binned_data_file <- file.path(comparison_dir, paste0(ccrix_cross, "_binned_concordance_data.csv"))
      write.csv(combined_binned, binned_data_file, row.names = FALSE)
      
      # Get all chromosomes
      chromosomes <- unique(combined_binned$chromosome_name)
      chromosomes <- sort(chromosomes)  # Sort chromosomes
      
      # Create chromosome-specific plots for this cross comparison
      for (chr in chromosomes) {
        # Filter data for this chromosome
        chr_data <- combined_binned %>%
          dplyr::filter(chromosome_name == chr, !is.na(pct_concordance))
        
        if (nrow(chr_data) > 0) {
          # Set up colors based on the labels in the data
          plot_colors <- c()
          
          # Add colors for parental strains (using consistent colors)
          for (label in unique(chr_data$label)) {
            if (label == paste0(strain1, "x", strain2)) {
              # This is the CCRIX cross
              plot_colors[label] <- "yellow"
            } else if (label == "CC001xCC001") {
              plot_colors[label] <- "red"
            } else if (label == "CC019xCC019") {
              plot_colors[label] <- "blue" 
            } else if (label == "CC040xCC040") {
              plot_colors[label] <- "green"
            } else {
              # Fallback color if we encounter an unexpected cross
              plot_colors[label] <- "purple"
            }
          }
          
          # Create comparison plot
          comp_plot <- ggplot(chr_data, 
                             aes(x = position_bin, y = pct_concordance, 
                                 color = label, group = label)) +
            geom_line(linewidth = 1) +
            geom_point(size = 1.5) +
            ylim(0, 100) +
            labs(
              title = paste0("Concordance Comparison - Chromosome ", chr),
              subtitle = paste0("CCRIX: ", strain1, "x", strain2, " vs Parent Strains"),
              x = "Chromosome Position (Mb)",
              y = "% Concordance",
              color = "Cross"
            ) +
            scale_color_manual(values = plot_colors) +
            theme_minimal() +
            theme(
              plot.title = element_text(hjust = 0.5, face = "bold"),
              plot.subtitle = element_text(hjust = 0.5)
            )
          
          # Save comparison plot as JPEG
          comp_plot_file <- file.path(comparison_dir, 
                                     paste0("concordance_chr", chr, ".jpg"))
          ggsave(comp_plot_file, comp_plot, width = 12, height = 8, device = "jpeg", dpi = 300)
          cat("Saved concordance comparison for chromosome", chr, "to:", comp_plot_file, "\n")
        }
      }
    }
  } else {
    cat("\nWarning: Either no CC or no CCRIX data was processed successfully.\n")
    cat("CC processed crosses:", length(cc_processed), "\n")
    cat("CCRIX processed crosses:", length(ccrix_processed), "\n")
  }
  
  cat("\n==================================================\n")
  cat("Concordance comparison analysis completed successfully!\n")
  cat("Output files are in:", output_dir, "\n")
  cat("==================================================\n")
}

# ============================================================================
# MAIN ANALYSIS FUNCTION
# ============================================================================

compare_cc_and_ccrix <- function(cc_results_dir, 
                               ccrix_results_dir,
                               genes_with_id_path,
                               similarity_threshold,
                               output_base_dir,
                               bin_size_mb = 5) {
  
  # Load gene coordinates
  cat("Loading gene coordinates...\n")
  gene_coords <- load_gene_coordinates(genes_with_id_path)
  
  # Load CC and CCRIX results
  cat("Loading CC analysis results...\n")
  cc_results <- load_cc_results(cc_results_dir)
  
  cat("Loading CCRIX G0 analysis results...\n")
  ccrix_results <- load_ccrix_results(ccrix_results_dir)
  
  # Create output directory if it doesn't exist
  if (!dir.exists(output_base_dir)) {
    dir.create(output_base_dir, recursive = TRUE)
    cat("Created output directory:", output_base_dir, "\n")
  }
  
  # Create concordance comparison visualizations
  create_concordance_comparisons(
    cc_results = cc_results,
    ccrix_results = ccrix_results,
    gene_coords = gene_coords,
    similarity_threshold = similarity_threshold,
    bin_size_mb = bin_size_mb,
    output_dir = output_base_dir
  )
  
  return(TRUE)
}

# ============================================================================
# RUN THE ANALYSIS
# ============================================================================

# Execute the analysis with the parameters defined at the top of the script
result <- compare_cc_and_ccrix(
  cc_results_dir = cc_results_dir,
  ccrix_results_dir = ccrix_results_dir,
  genes_with_id_path = genes_with_id_path,
  similarity_threshold = similarity_threshold,
  output_base_dir = output_base_dir,
  bin_size_mb = bin_size_mb
)
```

# END

# 

## 
